---
title: "Lotka-Volterra Models"
author: "Boris Steipe <boris.steipe@utoronto.ca>"
date: "2024-10-22"
output: 
  html_document:
#    number_sections: yes
#    toc: yes
---

<!--
This document is an "R markdown" document. 

There are two ways to work with this file:

A: Rendering to HTML
====================
Click on the "Knit" button at the top, to run the code contained in this file.
You will obtain a structured document in HTML format that opens in the
Web browser that is built into RStudio. 

B: Working Interactively
========================
You can change the code in the code "chunks" to try variations interactively.
In this case you have to have to (a) prepare the environment by clicking on the
"Run All Chunks Above" icon at the right-hand side of the chunk, (b) make your
changes, and (c) click on the green right-pointing triangle "Run Current Chunk"
to execute the changed code.

It is probably best to first render the document (Option A), study the
rendered result, and then go back to specific points for experimentation
(Option B) to understand the code better.

-->

```{css, echo = FALSE}
.striped tr:nth-child(even) {
  background: #eaf1ff;
}
.striped td, .striped th {
  padding: 5px;
}
.meta {
  font-size: 0.8em;
  color: #bbbbbb;
}
```

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))

if (! requireNamespace("here", quietly=TRUE)) {
  install.packages("here")
}

knitr::opts_chunk$set(echo = TRUE,
                      cache.path = 'tmp/')

# Load the .util.R file into the document's workspace.
# Note: all paths of files and files that are source()'d by files have to
# be given via here::here() since development uses the project home as 
# working directory but knitr uses the directory that contains the script.
# (This required far more debugging and adjustments than I care for ...)
source(here::here(".util.R"))

```

<small>Explorations of the Lotka-Volterra equation as coupled ODEs using the deSolve package and Euler's method.
(c) Boris Steipe 2022-2024.</small>

----

This Markdown document is a bit technical in its code details, but I hope accessible enough in its _results_. You don't really need to understand the code details - unless you want to design your own coupled differential equation exploration. But read along, and follow up with your gAI assistant if you have additional questions.

----


## Introduction

Combining change and resistance to change are hallmarks of life, as it evolves at the boundary between static order and chaotic dissipation. Models of change are usually expressed as differential equations: mathematical expressions that calculate the change of a "state" over time. A _state_ can be a feature like length or weight, the size of a population, or the mean of a distance, depending on the question at hand. Such state-change equations are typically non-linear and incorporate geometric growth terms (i.e. something increases by a percentage of the current value at each time step, rather than by a constant amount). 

Whenever two or more different entities interact, and their respective values appear in the same equation, we speak of a "coupled" differential equation, and even when these are simple systems, they are generally not solvable by analytic methods but need numeric methods (i.e. we can't solve them by applying a formula but need to compute change over time), and they can display intriguingly complex behaviour.

One particular model of two coupled non-linear differential equations is the Lotka-Volterra equations, named after Alexander Lotka and Vito Volterra who independently discovered their use to describe predator-prey interactions (cf. [Lotka-Volterra Equations (WP)](https://en.wikipedia.org/wiki/Lotka%E2%80%93Volterra_equations)). The basic equations are:

  $$\frac{dx}{dt} = ax - bxy$$
  $$\frac{dy}{dt} = gxy - hy$$
  
...with $x$ and $y$ representing the prey and predator populations and $a$,$b$,$g$,and $h$ being positive, real parameters.

In this exploration:

* We solve the equations and analyze the population numbers over time;
* we introduce time-series plots and phase space plots;
* we perform the calculation in a stepwise approximation (Euler's method)
    and explore the effect of random perturbations;
* we show that a systematic perturbation puts the system into an altered,
    but still stable configuration.
* we show how a simple three-species model evolves in a "strange" attractor
  basin and has chaotic behaviour;
* and we plot the trajectory in 3D.

## Initializations

### Packages

Check that required packages have been installed. Install if needed. Get information about the package contents"

```{r initializePackages}
# A package to compute ODEs ...
if (! requireNamespace("deSolve", quietly=TRUE)) {
  install.packages("deSolve")
}
# Package information:
#  library(help   = deSolve)     # basic information
#  browseVignettes("deSolve")    # available vignettes
#  data(package  = "deSolve")    # available datasets


# A package to support plotting. We need this later to plot lines with a colour
# gradient which we can't do in base R.
if (! requireNamespace("plotrix", quietly=TRUE)) {
  install.packages("plotrix")
}
# Package information:
#  library(help   = plotrix)     # basic information
#  data(package  = "plotrix")    # available datasets


# A package to support plotting 3D plots ...
if (! requireNamespace("plot3D", quietly = TRUE)) {
  install.packages("plot3D")
}
#  library(help   = plot3D)     # basic information
#  data(package  = "plot3D")    # available datasets
```


### Global Values

```{r globals}
# Color definitions for a consistent appearance of plots
PREYCOL <-  "#04ccc2"  # Color for prey in plots: "oceanic"
PREY2COL <- "#427deb"  # A second prey species: "seljuk blue"
PREDCOL <-  "#c41b65"  # Color for predators in plot: "bright rose"

# check:
# barplot(c(1, 1, 1), axes = FALSE, col = c(PREYCOL, PREY2COL, PREDCOL))
```


### Define Functions



####  ```LVmod()```  

A Lotka-Volterra Model for use in an ODE solver:

```{r Lvmod}
LVmod <- function(t, state, param) {
  #' LVmod - Lotka-Volterra Model
  #'
  #' @param t      num  A vector of time points
  #' @param state  num  A named vector with starting values for nPrey and nPred.
  #' @param param  num  A named vector containing the parameters of the model:
  #'                    preyGrowthRate, predationRate, predatorGrowthRate, and
  #'                    predatorLossRate.
  #'
  #' @return num        A list of population changes "dPrey" and "dPred".
  #'
  #' @details  This function is meant to be used by the deSolve::ode() ODE
  #'           solver to compute prey and predator populations over time.
  #'

  with(as.list(c(state, param)), {
    dPrey = (preyGrowthRate * nPrey) - (predationRate * nPrey * nPred)
    dPred = (predatorGrowthRate * nPrey * nPred) - (predatorLossRate * nPred)
    return(list(c(dPrey, dPred)))
  })
}
```

####  ```timePlot()```  

A convenience function to plot time-series of predator / prey populations:

```{r timePlot}

timePlot <- function(dat,
                     ix = 1,
                     iy = c(2, 3),
                     col = c(PREYCOL, PREDCOL),
                     main = "",
                     pMax = min(2000, nrow(dat))) {
  #' timePlot - Time-series Plot
  #'
  #' @param  dat  num  a numeric matrix that contains x-axis values (time) and
  #'           `       population values.
  #' @param  ix   int  column index of time values. Default: 1
  #' @param  iy   int  column indices of population values. Default: c(2, 3)
  #' @param  col  chr  color values. Default c(PREYCOL, PREDCOL).
  #' @param  main chr  plot title. Default "".
  #' @param  pMax num  The maximum number of points to plot. Some time-series
  #'                   can be quite long and the plots can take a long time
  #'                   to plot and update. But we generally do not need to
  #'                   plot at the full resolution of the time-series. However
  #'                   just plotting every n-th point or so creates plotting
  #'                   artefacts. Instead, we space points by distance along the
  #'                   plotted line. Default is the min() of 2000 points or
  #'                   the number of rows in the dataset.
  #'
  #' @return           NULL. The function plots a time series.
  #'
  #' @details          This is a standard time-series plot but written so that
  #'                   it can plot a variable number of columns, and use only
  #'                   a subset of equally spaced points. The column
  #'                   indices for the values to be plotted along the y-axis
  #'                   are passed in the variable iy. Axis labels are taken
  #'                   from the colnames of the matrix that contains the values
  #'                   to be plotted. First, an empty frame is plotted with
  #'                   enough space to accommodate the values, then the
  #'                   lines() function is used to plot the value of each
  #'                   column in turn. Finally, a legend is added.

  # empty plot frame
  plot(dat[, ix], dat[, iy[1]],
       type = "n",
       ylim = c(0, max(dat[, iy] * 1.3)),
       xlab = colnames(dat)[ix],
       ylab = paste(colnames(dat)[iy], collapse = ", "),
       main = main)

  # values
  for (i in 1:length(iy)) {
    sel <- selP(dat[,c(ix, iy[i])], pMax)
    lines(dat[sel, ix], dat[sel, iy[i]],
          type = "l",
          col = col[i])
  }

  # add a legend
  legend("topright",
         legend = colnames(dat)[iy],
         lty = rep(1, length(iy)),
         col = col,
         bty = "n")

  return(invisible(NULL))
}

```

#### ```selP()```

A convenience function to subset large datasets for plotting.

```{r selP}

selP <- function(x, N) {
  #' selP - select points for a line-plot
  #'
  #' @param x  num
  #' @param N  Number of points to be selected.
  #'
  #' @details  A vector of indices is produced that selects from a vector or
  #'           matrix, according to the Euclidian distance between two
  #'           consecutive elements.
  #'
  #' @return  a vector of N row-indices for subsetting x according to the
  #'          distance between consecutive rows. The first and last points
  #'          are always included.

  if (is.null(nrow(x))) { x <- as.matrix(x, byrow = TRUE) }
  l <- nrow(x)                                    # number of rows
  d <- sqrt(rowSums((x[2:l, ] - x[1:(l-1), ])^2)) # distances
  d <- c(0, d)                                    # add back first element

  # compute the cumulative sums, cut them into (N-1) intervals, get the
  # row indices and add the last element too.
  sel <- cumsum(d)
  sel <- cut(sel, breaks = (N-1), include.lowest=TRUE, labels = FALSE)
  sel <- c(which(! duplicated(sel)), l)

  return(sel)
}

```

#### ```phasePlot()```

A convenience function to plot trajectories in phase-space of a predator/prey time series.

```{r phasePlot}

phasePlot <- function(dat,
                      ix = 2,
                      iy = 3,
                      colMode = "ratio",
                      col = c(PREYCOL, PREDCOL),
                      transp = "44",
                      main = "",
                      pMax = min(2000, nrow(dat))) {
  #' phasePlot - Phase-space Plot
  #'
  #' @param  dat  num  a numeric matrix that contains population values
  #'                   in columns ix and iy.
  #' @param  ix   int  column index of first population. Default: 2
  #' @param  iy   int  column index of second population. Default: 3
  #' @param  colmode chr  "ratio": color by ratio of species1/species2
  #'                      (default); "length": color by length of the
  #'                      trajectory from first to last element.
  #' @param  col  chr  color values. Default c(PREYCOL, PREDCOL).
  #' @param  transp  hex  Transparency value. Added to plot colors.
  #' @param  main chr  plot title. Default "".
  #' @param  pMax num  The maximum number of points to plot. Some trajectories
  #'                   can be quite long and the plots can take a long time
  #'                   to plot and update. But we generally do not need to
  #'                   plot at the full resolution of the trajectory. However
  #'                   just plotting every n-th point or so creates plotting
  #'                   artefacts. Instead, we space points by distance along the
  #'                   plotted line. Default is the min() of 2000 points or
  #'                   the number of rows in the dataset.
  #'
  #' @return           NULL. The function plots a phase-space trajectory.
  #'
  #' @details          This is a standard phase-space plot which plots two
  #'                   populations from a Lotka-Volterra plot or similar against
  #'                   each other. However it uses a colour gradient to plot the
  #'                   trajectory. The colour gradient is
  #'                   determined along the log-ratio of the two populations.
  #'                   First, an empty frame is plotted, then the
  #'                   plotrix::color.scale.line() function plots the
  #'                   trajectory. A transparency value is added to the
  #'                   standard colors to show when lines overlap.

  # subset the plotMatrix
  sel <-selP(dat[ , c(ix, iy)], pMax)
  x <- dat[sel, ix]
  y <- dat[sel, iy]

  # construct a color vector for the trajectory, either by
  # log-ratio of x and y, or by length of the trajectory
  if (colMode == "ratio") {

    # 50 colors from col[1] to col[2]
    makeMyPal <- colorRampPalette(c(col[1], col[2]))
    myPal <- makeMyPal(50)


    # Fetch each row's color based on the log ratio of x and y
    lr <- log(x/y)                             # log ratio
    lr <- (lr - min(lr)) / (max(lr) - min(lr)) # normalize
    lr <- round(lr * 49) + 1                   # scale from 1 to 50
    myCol <- myPal[lr]                           # assign one of the 50 colours


  } else if (colMode == "length") {

    myCol <- hcl.colors(length(x), "Berlin")    # choose a palette 

  } else {
    stop(sprintf("Unknown or missing colMode \"%s\".", colMode))
  }
  
  # Add the transparency value
  myCol <- gsub("$", transp, myCol)


  # empty plot frame
  plot(x, y,
       type = "n",
       xlab = colnames(dat)[ix],
       ylab = colnames(dat)[iy],
       main = main)

  # plot the phase-space trajectory
  plotrix::color.scale.lines(x, y, col = myCol, lwd = 3)

  return(invisible(NULL))
}
```

## Lotka-Volterra model via ```deSolve::ode()```

Compute and plot a Lotka Volterra model with an ODE solver.

```{r deSolve}

# Define the parameters
Par <- c(preyGrowthRate = 1,
         predationRate = 0.2,
         predatorGrowthRate = 0.08,
         predatorLossRate = 0.5)

# Define the initial state populations
State <- c(nPrey = 5, nPred = 2)

# Initialize the time-axis
Time <- seq(0, 100, by = 0.01)

# Run the ODE solver and compute the time series
myTS <- as.data.frame(deSolve::ode(func  = LVmod,
                                           y     = State,
                                           parms = Par,
                                           times = Time))

timePlot(myTS, main = "Lotka-Volterra Model (ODE)")

phasePlot(myTS, main = "Phase-Space of Lotka-Volterra Model (ODE)")

```

##  Lotka-Volterra model via Euler's Method 

See also Green and Shou (2014).

Let us rewrite our code to perform exactly the same computation of the coupled differential equations of the Lotka-Volterra Model above (up to numeric error), but step-by-step, from a starting condition. This is a method published by Leonhard Euler in 1768, thus it is known as "Euler's Method". The ODE solver we used above has a number of modern methods available, which generally result in a much smaller error. But calculating the process "manually" as we do below, makes it easier to explore different functions for predator and prey population interactions.

To be more flexible in our explorations, we tease apart the computational steps more cleanly. For example, our parameters concern values that we need for running the function over time, parameters that concern only the behaviour of prey, and parameters that concern only the predators. We can abstract this out into three different sets of parameters.

```{r EulerParameters}

myPar <- list(nPreyStart = 5,
              nPredStart = 2,
              Time = 100,
              dt = 0.01)

myParPrey <- list(preyGrowthRate = 1,
                  predationRate = 0.2)

myParPred <- list(predatorGrowthRate = 0.08,
                  predatorLossRate = 0.5)
```

Next we define a function that does the actual calculation. It takes as parameters the values we defined above, and two functions! A function in R is an object as any other, it can be stored in lists, passed as a parameter, assigned to names etc. etc. Abstracting the core equations out of our computation makes it easy to change the functions later and experiment with different ways to compute interactions. We just call our computation with different functions and we don't need to copy the rest of the code again and again. This ensures that our computations remain consistent!

```{r EulerEngine}

LVEuler <- function(par = myPar,
                    parPrey = myParPrey,
                    parPred = myParPred,
                    fPrey,
                    fPred) {

  N <- par$Time / par$dt                         # Number of steps to compute
  outM <-  matrix(numeric(N * 3), ncol = 3)      # Set up the output matrix
  colnames(outM) <- c("time", "nPrey", "nPred")

  nPrey <- par$nPreyStart                        # Initialize the starting
  nPred <- par$nPredStart                        # values

  outM[1, "time"]  <- 0.0                        # Store the starting values
  outM[1, "nPrey"] <- nPrey
  outM[1, "nPred"] <- nPred

  # Iterate ...
  for (i in 2:N) {                               # for all time points

    # compute the population change
    dPrey = fPrey(nSelf = nPrey, nOther = nPred, par = parPrey) * par$dt
    dPred = fPred(nSelf = nPred, nOther = nPrey, par = parPred) * par$dt

    # compute the new populations
    newPrey <- nPrey + dPrey
    newPred <- nPred + dPred

    # store the results
    outM[i, ] <- c((i-1) * par$dt, newPrey, newPred)

    # prepare for the next iteration
    nPrey <- newPrey
    nPred <- newPred
  }

  # Done: return the result
  return(outM)
}

# Next we define the equations that compute the population change. We write
# them as functions so we can easily modify them.

# ... for Prey
fLVPrey <- function(nSelf, nOther, par) {
  dSelf <- (par$preyGrowthRate * nSelf) -
           (par$predationRate  * nSelf * nOther)
  return(dSelf)
}

# ... for Predators
fLVPred <- function(nSelf, nOther, par) {
  dSelf <- - (par$predatorLossRate   * nSelf) +
             (par$predatorGrowthRate * nSelf * nOther)
  return(dSelf)
}
```

With this "refactoring" of our code - separating the driving code and the functional equations - we clearly distinguish those parts of the computation that implement the actual model, and those parts that we only need to run the 
model along a trajectory. 

This is an example of an important objective in software engineering, called "**separation of concerns**".

Now we are ready to run the model. Since we are using the same parameters as for our ODE solver, we expect the time series to look the same.

```{r runLVEuler}

myTS <- LVEuler(fPrey = fLVPrey,  # Note: no parentheses! We are passing the
                fPred = fLVPred)  # function itself, not the result of
                                  # running the function. And: all other
                                  # parameters are passed via the default
                                  # values

# plot the results ...
timePlot(myTS, main = "Lotka-Volterra Model (Euler)")

# ... with predator / prey colors
phasePlot(myTS,
          main = "Phase-Space of Lotka-Volterra Model (Euler)")

# ... with colors spectrum from first (blue) to last (red) time point
phasePlot(myTS,
          main = "Phase-Space of Lotka-Volterra Model (Euler)",
          colMode = "length")

```


##  Adding Noise

Now we can reap the benefits of the increased abstraction of our code. It is easy add a bit of noise to the predation rate to see what effect noisy conditions have on the stability of the system. There are many ways to define "noise". Let's use a Gaussian model that increases or decreases the value of the predation rate by approximately +- 1/3 of its mean value. What does such added noise look like?


```{r rnorm}
myParPrey$predationRate   # 0.2 - the default value

nValues <- 10000
x <- rnorm(nValues,          # What does this distribution look like?
           mean = myParPrey$predationRate,
           sd = myParPrey$predationRate / 3)

head(x)  # print the first few actual values ...

# plot this as a histogram
hist(x, breaks = 50)
abline(v = myParPrey$predationRate, col = "#AA0000", lwd = 3)

# plot the actual values
myCols <- hcl.colors(nValues,             # Get the basic colors
                     palette = "Berlin")  
myCols <- gsub("$", "22", myCols)          # Make them transparent - set a value
                                           # of "22" in the "Alpha channel" at
                                           # the end of a color hex value, e.g
                                           # "#23D7CA" turns into  "#23D7CA22".
plot(x, ylim = c(-0.2, 0.7), pch = 21, col = myCols)                                         
```


###  Modifying our parameters 

```{r noisyParameters}

# Here is the function that computes prey populations, modified to add noise
# to :
fLVPreyStoch <- function(nSelf, nOther, par) {
  R <- par$predationRate
  R <- rnorm(1, mean = R, sd = R / 3)
  dSelf <- (par$preyGrowthRate * nSelf) - (R * nSelf * nOther)
  return(dSelf)
}

# plot
outS <- LVEuler(fPrey = fLVPreyStoch, fPred = fLVPred)
timePlot(outS, main = "Lotka-Volterra Model (Euler)")
phasePlot(outS, main = "Stochastic MOdel I")
```


Add more stochasticity: 33% of noise on the predator loss rate

```{r moreNoise}

fLVPredStoch <- function(nSelf, nOther, par) {
  R <- par$predatorLossRate
  R <- rnorm(1, mean = R, sd = R / 3)
  dSelf <- - ( R  * nSelf) + (par$predatorGrowthRate * nSelf * nOther)
  return(dSelf)
}

myTS <- LVEuler(fPrey = fLVPreyStoch, fPred = fLVPredStoch)
timePlot(myTS, main = "Lotka-Volterra Model (Euler)")
phasePlot(myTS, main = "Stochastic Model II")

```

Even more noise: 50% of noise on all rates ...

```{r evenMoreNoise}

fLVPreyStoch <- function(nSelf, nOther, par) {
  R1 <- par$predationRate
  R1 <- rnorm(1, mean = R1, sd = R1 / 2)
  R2 <- par$preyGrowthRate
  R2 <- rnorm(1, mean = R2, sd = R2 / 2)
  dSelf <- (R2 * nSelf) - (R2 * nSelf * nOther)
  return(dSelf)
}

fLVPredStoch <- function(nSelf, nOther, par) {
  R1 <- par$predatorLossRate
  R1 <- rnorm(1, mean = R1, sd = R1 / 2)
  R2 <- par$predatorGrowthRate
  R2 <- rnorm(1, mean = R2, sd = R2 / 2)
  dSelf <- - ( R1  * nSelf) + (R2 * nSelf * nOther)
  return(dSelf)
}

myTS <- LVEuler(fPrey = fLVPreyStoch, fPred = fLVPredStoch)
timePlot(myTS, main = "Lotka-Volterra Model (Euler)")

phasePlot(myTS, main = "Stochastic Model II")  # ... by ratio

phasePlot(myTS, main = "Stochastic Model II", colMode = "length") # ... by time
```


Let's see what happens if we make this trajectory longer:

```{r }
myPar$Time <- 500  # was 100
outS <- LVEuler(fPrey = fLVPreyStoch, fPred = fLVPredStoch)
phasePlot(outS, main = "Stochastic Model II", colMode = "length")
```

The trajectory becomes noisy, but on average, the behaviour is no different from the model without perturbations. In particular, we do not see that the model goes into an instable state, or becomes chaotic ...



##  Perturbations via deSolve::ode

What happens when we apply a momentary perturbation to the model?

```{r LVperturb}
LVperturb <- function(t, state, param) {

  with(as.list(c(state, param)), {
#    doPerturbation <- runif(1) < 0.005
    dPrey = (preyGrowthRate * nPrey) - (predationRate * nPrey * nPred)
    dPred = (predatorGrowthRate * nPrey * nPred) - (predatorLossRate * nPred)
    if (t > 29.5 && t < 33) {  # for a given time-window
      dPrey <- dPrey * 2       # multiply prey-change by two
      dPred =  dPred / 2       # divide predator-change by 2
    }
    return(list(c(dPrey, dPred)))
  })
}

Par <- c(preyGrowthRate = 1,
         predationRate = 0.2,
         predatorGrowthRate = 0.08,
         predatorLossRate = 0.5)
State <- c(nPrey = 5, nPred = 2)
Time <- seq(0, 100, by = 0.01)

# Run the ODE solver
out <- as.data.frame(deSolve::ode(func  = LVperturb,
                                  y     = State,
                                  parms = Par,
                                  times = Time))

timePlot(out, iy = c(2,3), main = "Lotka-Volterra Model (ODE)")

phasePlot(out, main = "Perturbation trajectory")

phasePlot(out, main = "Perturbation trajectory", colMode = "length")

```

The perturbation shifts the system out of its equilibrium, but it quickly swings back to a new, stable(!) regime.


##  Going 3D: Chaos, Finally

Two species are not enough to create instable, chaotic systems. This is reminiscient of physics, where two-body problems are generally solvable. However for three-body problems this is not the case. They are inherently unstable. So, let's add a third species into our model.

First we abstract out the basic equation. All our equations have an intrinsic term (growth rate or loss rate), and an interaction term (predation rate, an# predatorGrowth rate). But they are fundamentally the same type of equation: $\frac{dx}{dt} = ax - bxy$

We had switched the order of the two components of the equation for predators - but that is just the same as keeping the order and changing the sign on the parameters. We can abstract out the equation and just change our parameters accordingly.

### ```fDE3()```: difference equation for a three-species model.

```{r fDE3}

fDE3 <- function(p, n1, n2, n3) {
  dx = (p[1]*n1) - (p[2]*n1*n1) - (p[3]*n1*n2) - (p[4]*n1*n3)
  return(dx)
}

```

###```LV3D()``` - a three-species LV model that computes differences per time-step with ```fDE3()```.

```{r LV3D}
LV3D <- function(t, state, param) {

  with(as.list(c(state, param)), {
    dA = fDE3(param[1, ], nA, nB, nC)
    dB = fDE3(param[2, ], nB, nA, nC)
    dC = fDE3(param[3, ], nC, nA, nB)
    return(list(c(dA, dB, dC)))
  })
}

# The initial parameters for starting conditions: species A and B are prey
# species, species C is a predator.
State <- c(nA = 5, nB = 5, nC = 2)
Time <- seq(0, 100, by = 0.01)
col3 <- c(PREYCOL, PREY2COL, PREDCOL)   # colours

# The individual growth rates, and the pairwise interaction terms are as
# follows. The values were taken from Gilpin (1979).

Par <- matrix(numeric(12), nrow = 3)
Par[1,1] <-  1.0      # r1      Prey 1
Par[1,2] <-  0.001    # nA n1  n1n1 nAnA: a11
Par[1,3] <-  0.001    # nB n2  n1n2 nAnB: a12
Par[1,4] <-  0.01     # nC n3  n1n3 nAnC: a13

Par[2,1] <-  1.0      # r2      Prey 2
Par[2,2] <-  0.001    # nB n1  n1n1 nBnB: a22
Par[2,3] <-  0.0015   # nA n2  n1n2 nBnA: a21
Par[2,4] <-  0.001    # nC n3  n1n3 nBnC: a23

Par[3,1] <- -1.0      # r3      Predator
Par[3,2] <-  0.0      # nC n1  n1n1 nCnC: a33
Par[3,3] <- -0.005    # nA n2  n1n2 nCnA: a31
Par[3,4] <- -0.0005   # nB n3  n1n3 nCnB: a32
```


## Running the three-species model

Running the  model works just like for the 2D model ...

```{r run3Dmodel.1}
myTS <- as.data.frame(deSolve::ode(func=LV3D, y=State, parms=Par, times=Time))
timePlot(myTS, iy = c(2,3,4), main = "LV3D", col = col3 )

```

Intriguing ... Let us increase the simulation time

```{r run3Dmodel.2}
Time <- seq(0, 500, by = 0.01)
myTS <- as.data.frame(deSolve::ode(func=LV3D, y=State, parms=Par, times=Time))
timePlot(myTS, iy = c(2,3,4), main = "LV3D", col = col3 )
```

... Much longer:

```{r run3Dmodel.3}
# Patience: we have 500,000 steps to compute here ...
Time <- seq(0, 5000, by = 0.01)
myTS <- as.data.frame(deSolve::ode(func=LV3D, y=State, parms=Par, times=Time))
timePlot(myTS, iy = c(2,3,4), main = "LV3D", col = col3 )
```

What does the phase-space look like ? There are three 2D trajectories:

```{r run3Dmodel.4}

# Prey A (column 2) vs. the Predator (column 4)
phasePlot(myTS, ix=2, iy=4, main = "LV 3D", colMode = "length")

# Prey B (column 3) vs. the Predator (column 4)
phasePlot(myTS, ix=3, iy=4, main = "LV 3D", colMode = "length")

# Prey A (column 2) vs. Prey B (column 3)
phasePlot(myTS, ix=2, iy=3, main = "LV 3D", colMode = "length")
```

These are 2D projections of a 3D dataset. To get a better sense what the data actually looks like, we should plot this dataset in 3D:

The plotting functions we wrote above, have code that samples the number of points along the trajectory. For the lines3D function of the plot3D:: package, we should prepare the dataset and subset it. R has no problems to handle datasets with millions of rows - but one would not want to plot them directly.

```{r run3Dmodel.5}

N <- 3000                          # limit to N points along the trajectory

sel <- selP(myTS[ , c(2, 3, 4)], N) # subset according to the distance between
                                   # points on the trajectory

plot3D::lines3D(myTS[sel,2], myTS[sel,3], myTS[sel,4],
                colkey = FALSE,
                phi=20, theta=100)
```

Note that the color gradient that ```plot3D::lines3D()``` applies, colours values along the Z axis value, not along the time axis. Too bad. 


We can plot this as a 3D Stereoplot

```{r stereoPlot}

# dev.new()
opar <- par(mfcol = c(1,2),      # two images side-by-side
            mar   = c(1,1,1,1))  # reduce plot-margin widths

plot3D::lines3D(myTS[sel,2], myTS[sel,3], myTS[sel,4], # left-hand plot
                colkey = FALSE,
                phi=-3, theta=98)

plot3D::lines3D(myTS[sel,2], myTS[sel,3], myTS[sel,4], # right-hand plot
                colkey = FALSE,
                phi=-3, theta=102)   # rotate CCW by 5 degrees

par(opar)  #reset the plot parameters
```

The stereo plot clearly shows us that the trajectory does not fill phase space, but is confined to a particular region on the surface of a rounded tetragon. This is what is called a _limit cycle_, or _attractor_ of the trajectory: but there is a catch. The trajectory is not confined to a defined curve, but starts filling (at least parts of) the bounding surface. Thus geometrically speaking it lies somewhere _between_ a 1-dimensional curve and a 2-dimensional plane! This makes the object that is being traced out by our system's trajectory a _fractal_ - an object with a non-integer, "broken" (lat.: _fractus_: broken) dimension. 

Benoît Mandelbrot introduced the term _fractal_ in 1975, for structures that cannot be described by classical Euclidean geometry, where dimensions are whole numbers (1D, 2D, 3D... ). Instead, fractals can have dimensions that are non-integer.

And this in turn makes this structure a very special type of attractor: a "strange attractor", it is strange in the sense that it _attracts_ the trajectory in phase-space, but the trajectory does not settle into a precise, repetitive cycle.

ChatGPT describes this as follows:


>A strange attractor is a type of attractor that arises in dynamical systems exhibiting chaos. It is characterized by:
>
>* **Chaotic behavior**: The system is sensitive to initial conditions, meaning that nearby trajectories in phase space diverge exponentially, even though they remain bounded.
>
>* **Fractal structure**: Strange attractors typically have a fractal (non-integer) dimension, meaning that they exhibit self-similarity and complex, intricate structures when examined in detail.
>
>* **Attracting region**: Despite the chaotic nature of the system, trajectories are drawn to a certain bounded region of phase space rather than diverging to infinity.


Order and chaos, phase-space trajectories, fractals and strange attractors. These are some of the most intriguing ideas that arise in the study of dynamical systems.

That's all.

----

## Further Reading ##

This code has its origins in a blog entry in "Mind of a Markov Chain" (2010) that is no longer public. It was preserved in the R-Bloggers aggregation. (Shout out to [R-Bloggers]( https://www.r-bloggers.com/). The link is [here]( https://www.r-bloggers.com/2010/03/lotka-volterra-model%C2%A0%C2%A0intro/) The author of the original post is "apeescape". 

Apparently that post (uncredited) inspired Green and Shou (2014) to a very much more detailed exposition in a Methods in Molecular Biology chapter: ["Modeling community population dynamics with the open-source language R."]( https://pubmed.ncbi.nlm.nih.gov/24838889/)  (pdf available.). 

The three-species model with a strange attractor was published by Michael Gilpin (1979) [Spiral Chaos in a Predator Prey Model. The American Naturalist. 113(2): 306-308]( https://www.jstor.org/stable/pdf/2460209.pdf).

<p>&nbsp;
<p>

----

<p>&nbsp;
<p>

<div class="meta">
#### Meta ####

* Version 2024.1.0
* Code derived in part from `LotkaVolterra-Model.R` written for CSB195 2022 and 2023.

</div>


<!-- [END] -->
