---
title: "The Edge of Chaos"
author: "Boris Steipe <boris.steipe@utoronto.ca>"
date: "2024-10-22"
output: 
  html_document:
#    number_sections: yes
#    toc: yes
---

<!--
This document is an "R markdown" document. 

There are two ways to work with this file:

A: Rendering to HTML
====================
Click on the "Knit" button at the top, to run the code contained in this file.
You will obtain a structured document in HTML format that opens in the
Web browser that is built into RStudio. 

B: Working Interactively
========================
You can change the code in the code "chunks" to try variations interactively.
In this case you have to have to (a) prepare the environment by clicking on the
"Run All Chunks Above" icon at the right-hand side of the chunk, (b) make your
changes, and (c) click on the green right-pointing triangle "Run Current Chunk"
to execute the changed code.

It is probably best to first render the document (Option A), study the
rendered result, and then go back to specific points for experimentation
(Option B) to understand the code better.

-->

```{css, echo = FALSE}
.striped tr:nth-child(even) {
  background: #eaf1ff;
}
.striped td, .striped th {
  padding: 5px;
}
.meta {
  font-size: 0.8em;
  color: #bbbbbb;
}
```

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))

if (! requireNamespace("here", quietly=TRUE)) {
  install.packages("here")
}

knitr::opts_chunk$set(echo = TRUE,
                      cache.path = 'tmp/')

# Load the .util.R file into the document's workspace.
# Note: all paths of files and files that are source()'d by files have to
# be given via here::here() since development uses the project home as 
# working directory but knitr uses the directory that contains the script.
# (This required far more debugging and adjustments than I care for ...)
source(here::here(".util.R"))

```

<small>Chaotic Sytems explorations: Logistic map, Hénon map, and the double pendulum.
(c) Boris Steipe 2022-2024.</small>

----

This Markdown document is a bit technical in its code details, but I hope accessible enough in its _results_. You don't really need to understand the code details - unless you want to design your own coupled differential equation exploration. But read along, and follow up with your gAI assistant if you have additional questions.

----


## Introduction

Life exists at the delicate intersection of order and chaos—a dynamic edge where complexity thrives. To survive and evolve, living systems must balance the conflicting demands of resisting dissipation yet enabling adaptation. This balance requires state-spaces with unbounded complexity, allowing for stable existence, yet supporting variation and change wherever necessary. In this Markdown document, we  explore the foundational concepts of chaos theory through the lens of the three classic examples: logistic map, the Hénon map, and the double pendulum. Through writing the actual code that underlies these systems, and generating a number of diverse visualizations, we illuminate what chaotic systems are, and introduce essential vocabulary.

In this exploration:

* We compute the logisitc map and show how its image is generated from the behaviour of its function values at different choices of a parameter, and that the parameters can lead to constant as well as chaotic behaviour;
* we compute the Hénon map and zoom in 10,000-fold to illustrate its attractor - a non-geometric mathematical object that has a dimensionality somewhere between 1D and 2D;
* we model and solve the equations that govern a double pendulum, demonstarte its behaviour in cyclical and chaotic domains, and introduce three ways of looking at the trajectory: in angular space, in phase space, and by analyzing the power spectrum;
* and we plot the corresponding plots all together into the same window.

## Initializations

### Packages

Check that required packages have been installed. Install if needed. Get information about the package contents.

```{r initializePackages}
# A package to compute ODEs ...
if (! requireNamespace("deSolve", quietly=TRUE)) {
  install.packages("deSolve")
}
# Package information:
#  library(help   = deSolve)     # basic information
#  browseVignettes("deSolve")    # available vignettes
#  data(package  = "deSolve")    # available datasets


# A package to support plotting. We need this later to plot lines with a colour
# gradient which we can't do in base R.
if (! requireNamespace("plotrix", quietly=TRUE)) {
  install.packages("plotrix")
}
# Package information:
#  library(help   = plotrix)     # basic information
#  data(package  = "plotrix")    # available datasets

```


## The Logistic Map

The logistic map is a simple recurrence relation:
$$x_{n+1} = r * x_n * (1 − x_n)$$
Where:     
$x_n$:    the current state (between 0 and 1).<br/>
$r$:      an arbitrarily chosen parameter (typically between 0 and 4). It turns out that $r$ controls the behavior of the map.<br/>
$x_{n+1}$:  is the next state. 

N.b. A number of related terms describe ways to associate one "thing" with another: "relation", "map", "equation", "transformation", "operator", ... . The term "relation" in "recurrence relation" can  be a bit misleading: here it refers to an _equation_ that defines a  sequence in a recursive manner. The term "map" refers to the function's ability to "map" or "send" one value (from the "domain") to another value (in the "codomain"). This is common terminology in the study of dynamical systems. For the logistic map, the iteration "maps" each value in the interval [0,1] to some other value in [0,1].   

Let's examine the behavior of this relation:

### Define the basic function

####  ```runLogistic()```  

```{r runLogistic}
runLogistic <- function(x1, r, N) {
  # This function iterates the Logistic Map from the starting value
  # x1 for N iterations using the parameter r. It returns every value
  # x that is produced, in a vector.

  x <- numeric(N)                       # Initialize the vector of results
  x[1] <- x1                            # x1 is the first value
  for (n in 1:(N - 1)) {                # for N steps ...
    x[n+1] <- r * x[n] * (1.0 - x[n])   #  ... apply the relation.
  }
  return(x)
}
```


Different starting values `x1` converge to the same value quite quickly ...

``` {r converge}
N <- 50
plot(1:N, runLogistic(x1 = 0.1, r = 0.99, N = N),  # First line
     type = "l", xlim = c(0, N), ylim <- c(0, 0.6),
     main = "Logistic map: convergence",
     xlab = "number of iterations",
     ylab = "x",
     col = "#9f3352")

for (x1 in seq(0.1, 0.6, by = 0.1)) {               # A sequence of starting values
                                                    # Add more lines
  points((1:N), runLogistic(x1, 0.99, N), type = "l", col = "#9f3352")
}
```

All of these iterations converge quickly to the same $x$ value; that's not what makes the Logistic Map interesting. The relation behaves quite remarkably differently for different choices of $r$ !


``` {r divergence.1}
# Define a set of colors from a gradient
myCols <- colorRampPalette(c("#e1e6ff", "#D2BED8", "#9f3352"))(10)

N <- 50
x1 <- 0.25   # This is not a particularly special value

plot(1:N, runLogistic(x1 = x1, r = 0.1, N = N),
     pch = 20, xlim = c(0, N), ylim = c(0, 1),
     main = "Logistic map: Dependence on r",
     xlab = "number of iterations",
     ylab = "x",
     col = myCols[1])
#                         Watch what happens for different values of r:
points((1:N), runLogistic(x1 = x1, r = 0.999, N = N),  # r just below 1
       pch = 20, col = myCols[2])
points((1:N), runLogistic(x1 = x1, r = 1.1, N = N),    # r = 1.1
       pch = 20, col = myCols[3])
points((1:N), runLogistic(x1 = x1, r = 2.0, N = N),    # 2.0
       pch = 20, col = myCols[4])
points((1:N), runLogistic(x1 = x1, r = 2.9, N = N),    # 2.9
       pch = 20, col = myCols[5])
points((1:N), runLogistic(x1 = x1, r = 3.2, N = N),
       pch = 20, col = myCols[6])
points((1:N), runLogistic(x1 = x1, r = 3.4, N = N),
       pch = 20, col = myCols[7])
points((1:N), runLogistic(x1 = x1, r = 3.5, N = N),
       pch = 20, col = myCols[8])
points((1:N), runLogistic(x1 = x1, r = 3.572, N = N),
       pch = 20, col = myCols[9])
points((1:N), runLogistic(x1 = x1, r = 3.67, N = N),
       pch = 20, col = myCols[10])
```

For lower values of $r$ we converge to one $x$-value, but for higher values the function starts oscillating between two $x$-values, then four, then more.

There is a lot of overlap, so let's look again at the behaviour of the last six values of $r$ - we can draw them separately for clarity.

``` {r divergence.2}

plot(1:N, runLogistic(x1 = x1, r = 2.9, N = N),
     pch = 20, xlim = c(0, N), ylim = c(0, 6),
     main = "Logistic map: Dependence on r",
     xlab = "number of iterations",
     ylab = "x",
     yaxt = "n",
     col = myCols[5])
abline(h = 1:5)
points((1:N), runLogistic(x1 = x1, r = 3.2, N = N) + 1,
       pch = 20, col = myCols[6])
points((1:N), runLogistic(x1 = x1, r = 3.4, N = N) + 2,
       pch = 20, col = myCols[7])
points((1:N), runLogistic(x1 = x1, r = 3.5, N = N) + 3,
       pch = 20, col = myCols[8])
points((1:N), runLogistic(x1 = x1, r = 3.572, N = N) + 4,
       pch = 20, col = myCols[9])
points((1:N), runLogistic(x1 = x1, r = 3.67, N = N) + 5,
       pch = 20, col = myCols[10])
text(rep(5, 6), (0:5) + 0.1,
     labels = sprintf("r = %.3f",c(2.9, 3.2, 3.4, 3.5, 3.572, 3.67)))
```

At $r = 2.9$, instead of quickly moving to a converging value, we see some oscillation in $x$ values - but they do converge. However, at $r = 3.2$, we get divergence. At $r = 3.4$, there is a noticeable oscillation superimposed on the diverging values ... which gets stronger at $r = 3.5$, and at $r = 3.572$, we get oscillations on top of the oscillations.

Let's examine the distribution of the x values at  a slightly higher value of $r = 3.67$, and lets compute a large number $x$ values:


``` {r chaoticDomain}

x <- runLogistic(x1 = x1, r = 3.67, N = 1000000) # a million
h <- hist(x, breaks = 150, plot = FALSE)         # get the histogram values

# define 50 colors from the hcl palette from light to dark
myPal <- rev(c(hcl.colors(49), "#ffffff"))

# assign a color from the palette to each histogram bin, 
# depending on the number of observations in that bin
myCols <- myPal[round(1 + ((h$counts / max(h$counts)) * 49))]

# plot the histogram
hist(x, breaks = 150, col = myCols, border = myCols, xlim = c(0, 1))
```

You see that there are two bounding values for $x$, but that the space is pretty much filled. In fact, among the 1,000,000 values we have `r format(length(unique(x)), big.mark=",", scientific=FALSE)` unique values! There is no repetition.

To explore this behavior of $r$ across its entire range, we can produce many very fine-grained histograms of $x$ values, stack them vertically and plot them side by side. A single stack looks like this:

``` {r oneStack}

plot(rep(3.67, length(h$mids)), h$mids, type = "n")
points(rep(3.67, length(h$mids)), h$mids, pch = 15, col = myCols)
abline(v=3.67, col="#AAAADD")

```


###  Logistic Map - Supporting functions

Actually plotting the "Logistic Map" requires a few supporting functions, if you are curious, examine the code below to learn how things are done. Otherwise, skip to the next section where we plot the map.

``` {r functions.2}

logisticIteration <- function(r, xMin, xMax, x, nInit, nRun, nPoints) {
  # Compute one evolution of logistic map equation for a given r at some
  # x, by default 0.25.
  # break() when a value is repeated and multiply resulting histogram
  # to give correct overall number of trials
  IMAXMATCH <- 200  # tradeoff between early breaks (good) and overhead of
                    # finding matches (bad)

  br <- seq(xMin, xMax, length.out = (nPoints + 1))

  for (i in 1:nInit) {  # burn in from starting value
    x <- r * x * (1.0 - x)   # Logistic Map equation
  }
  v <- numeric(nRun)
  v[1] <- x
  for (i in 2:nRun) {
    x <- v[i-1]
    v[i] <- r * x * (1.0 - x)
    # if i > IMAXMATCH, no point in matching, there are likely going to be
    # very many unique values, so its faster to just compute them out.
    # Otherwise: match(), and break() if the values cycle.
    if (i < IMAXMATCH && !is.na(match(v[i], v[1:(i-1)]))) {
      break()
    }
  }
  v <- v[1:i]                       # use only computed values
  v <- v[(v >= xMin) & (v <= xMax)] # discard values outside domain, otherwise
                                    # hist() will complain
  h <- hist(v[1:i], breaks = br, plot = FALSE, warn.unused = FALSE)$counts
  h <- h * (nRun / i)
  return(h)
}


logisticImage <- function(rMin, rMax,
                          xMin, xMax,
                          nR, nX, x0,
                          nInit, nRun, showPbar) {
  # Helper function: iterate the map function many times to draw
  # an image of the logistic map in the specified region.
  img <- matrix(numeric(nR * nX), nrow = nX)
  vR <- seq(rMin, rMax, length.out = nR)

  for (i in 1:nR) {
    if (showPbar) {
      pBar(i, nR) # progress bar. Good for interactive explorations, not good
                  # in a Markdown document 
    }
    img[ ,i] <- logisticIteration(r = vR[i],
                                  xMin = xMin, xMax = xMax,
                                  x = x0,
                                  nInit = nInit, nRun = nRun,
                                  nPoints = nX)
  }
  return(img)
}


getGrid <- function(i, a, n) {
  # Helper function: compute grid points
  x <- seq(i, a, length.out = (n + 1))
  x <- x + ((x[2] - x[1]) / 2)
  return( x[-length(x)] )
}


plotLogisticMap <- function(rMin  = 2.5,
                            rMax  = 4.0,
                            xMin  = 0.0,
                            xMax  = 1.0,
                            nR    = 300,
                            nX    = 200,
                            x0    = 0.25,
                            nInit = 1000,
                            nRun  = 100000,
                            pal   = c(hcl.colors(63), "#FFFFFF"),
                            zMin  = 0.97,
                            showPbar = TRUE) {
  # Helper function: Plot the logistic map given the parameters
  img <- logisticImage(rMin = rMin, rMax = rMax,
                       xMin = xMin, xMax = xMax,
                       nR = nR, nX = nX, x0 = x0,
                       nInit = nInit, nRun = nRun,
                       showPbar)

  img <- t(1 - (img / max(img)))
  img <- ifelse(img < zMin, zMin, img)

  image(x = getGrid(rMin, rMax, nR),   # This determines the number of pixels
        y = getGrid(xMin, xMax, nX),   # that are being computed
        z = img,
        zlim = c(zMin, 1.0),
        col = pal,
        xlab = expression(italic("r")),
        ylab = expression(italic("x")),
        cex.axis = 0.8,
        main = sprintf("Logistic map between %3.2f and %3.2f", rMin, rMax),
        useRaster = TRUE)

}

```


### Visualizing the Logistic Map

We plot a (rather low resolution) logistic map, to orient ourselves.

``` {r logisticMapLowRes}

# Low resolution plot of a large domain of r.
plotLogisticMap(nR = 400, nX = 400, zMin = 0.97, showPbar = FALSE)

# Define a tiny region of the map to zoom in to:
rI <- 3.562
rA <- 3.575
xI <- 0.888
xA <- 0.894

rect (rI, xI, rA, xA,  border = "#CC0000")

```

This is the famous and intriguing structure of the logistic map. What I find most remarkable is the sense of overlapping, distinct domains - which is rather odd when you realize that there is nothing actually there that can overlap, it is just the propensity of certain numbers to behave in a certain way under this recurrence relation. Or, put differently: whenever we see structure in a pattern, that means in some sense adjacent elements are governed by a higher-order principle. But the $r$ values we are plotting here are just adjacent numbers, and the process is chaotic. This really illustrates the point that chaotic trajectories are not random.

I have defined a tiny, tiny region near the second bifurcation of the upper main branch and drawn a red rectangle around it. It's quite hard to see at this resolution. But let's zoom in:


``` {r logisticMapHiRes}

  #  Compute four million pixels (2000 * 2000).
  plotLogisticMap(rMin = rI, rMax = rA, xMin = xI, xMax = xA,
                  nR = 2000, nX = 2000, zMin = 0.995, showPbar = FALSE)

```

To actually see the plot at full resolution, and appreciate its intricate structure, click on "Show in a new window" in the markdown plot window when it is done and look at it on a large monitor.

You might think this is the same map as before - but it's not: it's a tiny region of the original map. This is what we mean when we say that these mathematical structurs can be self-similar at different scales. If we would make the interval even smaller, we would get similar structures. The $x$ values can lie arbitrarily close to each other. The relation in this domain is exquisitely sensitive to the values of $r$. 

**"Sensitive to initial conditions" - that is the definition of chaotic behaviour.**


## The Hénon map

The Hénon map is another "classical" chaotic system defined by the following _two_ coupled recurrence relations:

$$x_{n+1} = y_n +1 - a* x_n^2$$
$$y_{n+1} = b* x_n$$
We can look at the map's behaviour, to introduce the ideas of a "phase space" and an "attractor" in phase space. A _phase space plot_ plots one state-variable against the other state variable. An _attractor_ is a region of state space into which the points of a systems trajectory fall.

We first define the function that computes the Henon map:

``` {r henonFunction}
henonMap <- function(a, b, N, x0 = 0, y0 = 0) {
  # Compute the Henon map for parameters a and b, for N points, from
  # starting values x0 and y0. Then rescale the map into the unit square
  # (from (0,0) to (1,1) to make selecting and zooming easier.)

  x <- numeric(N)
  y <- numeric(N)
  x[1] <- x0
  y[1] <- y0
  for (n in 1:(N-1)) {  # iterate N times
    x[n+1] <- y[n] + 1 - a * x[n]^2
    y[n+1] <- b * x[n]
  }

  x <- (x - min(x)) / (max(x) - min(x))
  y <- (y - min(y)) / (max(y) - min(y))

  return(list(x = x, y = y))
}
```

We also define some small helper functions:

``` {r henonUtilities}

trimPoints <- function(map, box) {
  # Remove points from the list map: map $x and map$y that fall
  # outside the coordinates of box$xI, box$xA, box$yI, and box$yA. This is
  # useful to zoom in to large trajectories, we don't have to be slowed
  # down by the large number of points outside our zoomed region.

  sel <- map$x >= box$xI &
         map$x <= box$xA &
         map$y >= box$yI &
         map$y <= box$yA

  map$x <- map$x[sel]
  map$y <- map$y[sel]

  return(map)
}


sqBox <- function(x, y, wid) {
  # Utility function to define the coordinates of a square
  # box given the centre coordinates and the width.

  box <- list()
  box$xI <- x - (wid / 2)
  box$xA <- x + (wid / 2)
  box$yI <- y - (wid / 2)
  box$yA <- y + (wid / 2)

  return(box)
}
```

Compute the basic map.

``` {r computeMap}

# Parameters:
a <- 1.4
b <- 0.3
N <- 10000  # ten thousand iterations.
x0 <- 0     # initial condition for x
y0 <- 0     # initial condition for y

myMap <- henonMap(a, b, N, x0, y0)


```

Visualize the values of $x$ and $y$ over the length of the trajectory.

``` {r mapValues}


oPar <- par(mfrow = c(2,1), mar = c(4, 4, 1, 1))
plot(myMap$x[1:1000], type="l", col="#6688FF77")
plot(myMap$y[1:1000], type="l", col="#44CC6688")
par(oPar)

```

Individually, the $x$ and $y$ values appear to lie on a very noisy trajectory ...

... but things get interesting when we plot $x$-values against $y$-values, i.e. we produce a phase space plot. 

``` {r mapPhaseSpace, fig.asp = 1.0}

plot(myMap$x, myMap$y, pch=".", asp = 1.0)

# Define a region to zoom in on and draw a box around that region
boxCols <- hcl.colors(7, palette = "RdPu")[2:5]
box <- sqBox(x = 0.945, y = 0.585, wid = 0.1)
rect(box$xI, box$yI, box$xA, box$yA, border=boxCols[1])
```

This is the iconic shape of the Hénon map (at parameters $a = 1.4$ and $b = 0.3$). I have drawn a box around a region that we can zoom into in the next plot, at 10-times magnification. But we will recompute the map with many more points to allow us to zoom in.


``` {r zoom1, fig.asp = 1.0}

# Recompute the map with four million points
a <- 1.4
b <- 0.3
N <- 4000000  # Four million. It's a lot.
x0 <- 0       # initial condition for x
y0 <- 0       # initial condition for y

myMap <- henonMap(a, b, N, x0, y0)


# Trim points that are no longer visible and plot the rest
myMap <- trimPoints(myMap, box)
plot(myMap$x, myMap$y,
     pch=".",
     xlim = c(box$xI, box$xA),
     ylim = c(box$yI, box$yA),
     asp = 1.0)
rect(box$xI, box$yI, box$xA, box$yA, border=boxCols[1], lwd = 3)

box <- sqBox(x = 0.9228, y = 0.6245, wid = 0.01)
rect(box$xI, box$yI, box$xA, box$yA, border=boxCols[2])

```

10-fold magnification and I have drawn another rectangle for another magniifcation step.


``` {r zoom 2, fig.asp = 1.0}

# ... trim and plot
myMap <- trimPoints(myMap, box)
plot(myMap$x, myMap$y,
     pch=".",
     xlim = c(box$xI, box$xA),
     ylim = c(box$yI, box$yA),
     asp = 1.0)
rect(box$xI, box$yI, box$xA, box$yA, border=boxCols[2], lwd = 3)

# Zoom in even more
box <- sqBox(x = 0.92445, y = 0.62190, wid = 0.001)
rect(box$xI, box$yI, box$xA, box$yA, border=boxCols[3])
```

Even though the millions of points in our first plot made this structure appear like a closed curve, at this 100-fold magnification we can start to get the sense that we are actually looking at a point cloud.

Enhance ...


``` {r zoom3, fig.asp = 1.0}

# ... trim and plot
myMap <- trimPoints(myMap, box)
plot(myMap$x, myMap$y,
     pch=".",
     xlim = c(box$xI, box$xA),
     ylim = c(box$yI, box$yA),
     asp = 1.0)
rect(box$xI, box$yI, box$xA, box$yA, border=boxCols[3], lwd = 3)

# Zoom in one more time
box <- sqBox(x = 0.924628, y = 0.621712, wid = 0.0001)
rect(box$xI, box$yI, box$xA, box$yA, border=boxCols[4])
```


This is one-thousand fold. What you see here is an area that is smaller than a single pixel in our original plot. It is still visibly structured.

Let's go deeper ...


``` {r zoom4, fig.asp = 1.0}

# ... trim and plot
myMap <- trimPoints(myMap, box)
plot(myMap$x, myMap$y,
     pch=".",
     xlim = c(box$xI, box$xA),
     ylim = c(box$yI, box$yA),
     asp = 1.0)
rect(box$xI, box$yI, box$xA, box$yA, border=boxCols[4], lwd = 3)
```

Only a few points to show here - but we can clearly see that there is still remaining structure that would be resolved if we zoom even deeper.

Although the trajectory appeared to be a solid curve at lower magnification, it splits up into finer and finer structures as we zoom in, and this continues without bound. Curiously, the actual points of the trajectory do not lie next to each other, they just fall onto regions of a subspace that is nestled somehow in the curvy structures we have seen - though you could never say _exactly_ where it is, other than it exists at the location of the sampled points themselves. In that sense, the Hénon attractor is not actually a geometric object.

If you would like to explore this feature of strange attractors more deeply, here is a short conversation I had with ChatGPT on [the nature of the Hénon map](https://chatgpt.com/share/6717c6fc-df58-8009-ac77-9ad1f4385215).

If you would like to play with this some more, you could write a function that captures
x and y coordinates from your cursor, and then adds more points that fall into your box. So for the next zoom level, you might compute the map for ten million points, then a hundred million, etc. your computation may need to run for a while, but since you would be discarding most of the results, you will not overload your computer's memory. You could ask ChatGPT how to write such code that refills the visible points of the trajectory.



## The Double Pendulum[^1]

The double pendulum is a simple dynamic system with a direct physical interpretation. The equations governing its motion are a bit involved, but the system itself is easy to understand: it's just one pendulum attached to the end of another. The double pendulum is a classic example of a chaotic system in physics. When you set it into motion, it is nearly impossible to predict its behavior after a short amount of time, even though it is completely deterministic.

Physical Interpretation
: Imagine a pendulum (the "upper pendulum") hanging from a fixed point. Now, imagine attaching a second pendulum (the "lower pendulum") to the end of the first one. Even this simple mechanical system will exhibit chaotic behavior for certain initial conditions. Here is a video [Double Pendulum - MIT TSG](https://www.youtube.com/watch?v=czLIj-4suOk)

Biological Interpretation
: Chaotic dynamics can be found in various biological systems. One classic example is the beating of the human heart. Under certain conditions, the heart can enter chaotic rhythms (like fibrillation). While the heart's electrical system is vastly more complicated than a double pendulum, the idea is similar: a deterministic system that can exhibit unpredictable behavior. The double pendulum is just a metaphor in this case, but the dynamics we get from the interacting oscillators in the heart can look very similar to such a simple system. The we can ask: what aspects of the actual heartbeat can be mapped to the simple physical (or computational) model.

[^1]: Note: This example code was originally contributed by ChatGPT-4. The general layout of the input parameters and how to set up the function for the coupled differential equations and solve them with `desolve::ode()` was entirely correct - and that is very helpful. However, the details of the (complicated) first-order differential equations were wrong, even though they used the right elements of lengths, masses, and angular velocities, and `desolve::ode()` converged on solutions. The correct equations were taken from code in a nice interactive Web page at [myPhysicsLab](https://www.myphysicslab.com/pendulum/double-pendulum-en.html) and the Javascript source at [their GitHub page](https://github.com/myphysicslab/myphysicslab/blob/master/src/sims/pendulum/DoublePendulumSim).


## Define functions ...

### ```doublePendulum()``` - the function that computes the trajectory

``` {r douplePendulum}
doublePendulum <- function(t,        # timepoint
                           state,    # current values of theta and omega
                           parm) {   # lenghts and masses
  # This function computes the state of a double pendulum at timepoint t
  # given the two angles from the vertical, the angular velocities, and the
  # parameters for length and mass of both pendulums. deSolve::ode() solves this
  # equation at the requested timepoints to compute the trajectories for the two
  # pendulums over time.

  g <- 9.81  # (earth standard surface gravitational constant: 9.81 m/s^2)

  # Initial conditions
  th1 <- state["th1"]  # angle with the vertical of first pendulum
  om1 <- state["om1"]  # angular velocity of first pendulum (rad/sec)
  th2 <- state["th2"]
  om2 <- state["om2"]

  # Parameters
  L1 <-   parm[["L1"]]    # length of first pendulum
  m1 <-   parm[["m1"]]    # mass of first pendulum
  L2 <-   parm[["L2"]]
  m2 <-   parm[["m2"]]

  dy1 <- om1

  dy2 <- -g*(2*m1+m2)*sin(th1)
  dy2 <- dy2 - g*m2*sin(th1-2*th2)
  dy2 <- dy2 - 2*m2*om2*om2*L2*sin(th1-th2)
  dy2 <- dy2 - m2*om1*om1*L1*sin(2*(th1-th2))
  dy2 <- dy2/(L1*(2*m1+m2-m2*cos(2*(th1-th2))))

  dy3 <- om2

  dy4 <- (m1+m2)*om1*om1*L1
  dy4 <- dy4 + g*(m1+m2)*cos(th1)
  dy4 <- dy4 + m2*om2*om2*L2*cos(th1-th2)
  dy4 <- dy4*2*sin(th1-th2)
  dy4 <- dy4/(L2*(2*m1+m2-m2*cos(2*(th1-th2))))

  return(list(c(dy1, dy2, dy3, dy4)))
}
```


### Helper functions `deg2rad()` and `wrapTo2pi()`

``` {r helperFunctions}
deg2rad <- function(deg) {
  # convert degrees to radians
  return((deg / 180) * pi)
}

wrapTo2pi <- function(tra) {
  # Wrap theta values to the interval (-2pi, 2pi)
  tra[ , "th1"] <- tra[ , "th1"] %% (2 * pi)
  tra[ , "th2"] <- tra[ , "th2"] %% (2 * pi)
  return(tra)
}
```



### `plotPendulum()` - display various aspects of the trajectory


This function does a lot of work - analyzing the trajectories and displaying them in three separate sections of the plot-window. These combined plots are an example of a somewhat advanced level of data handling and visualization. You can certainly come back to this for your own visualization needs ...

``` {r plotPendulum}
plotPendulum <- function(tra, lwdTra = 1.0, lwdPhase = 2.0) {
  # plot of the double pendulum trajectory
  th1Col <- "#F085C5"
  th2Col <- "#9CA7EB"
  myCols <- colorRampPalette(
    c("#6d7c8a","#8d9199","#9c99a1","#a87d86","#c96772"))(nrow(tra))

  oPar <- par(mfrow = c(1, 2))

  # Define a layout for the plots: the next three plots go into the
  # sections defined in the matrix.
  myMat <- matrix(c(1, 2, 1, 2, 1, 3), ncol=2, byrow = TRUE)

  # Set the layout
  layout(myMat, widths=c(1,1), heights=c(1,1))

  # Next, we need a bit of additional logic to (mostly) keep trajectories
  # from breaking around 0 radians - since 0 and 2pi are identical points.
  # First we rotate the vertical axis from pointing down at th == 0, to
  # point down at th == pi. Then we take the values modulo 2pi. Finally
  # we plot the trajectories separately in sections, separated by breakpoints
  # the trajectory crosses the vertical. If we don't do that, we get a vertical
  # (or horizontal) line through our plot.

  # Rotate the trajectories: add pi to all theta values.
  tra[, "th1"] <- tra[, "th1"] + pi
  tra[, "th2"] <- tra[, "th2"] + pi

  # Take theta values modulo 2 pi
  tra[, "th1"] <- tra[, "th1"] %% (2 * pi)
  tra[, "th2"] <- tra[, "th2"] %% (2 * pi)

  # plot in sections: identify breaks in th1 and th2. The break point is
  # the first point on the next section
  idxBrkTh1 <- integer()
  idxBrkTh2 <- integer()
  for (i in 1:(nrow(tra) - 1)) {
    if (abs(tra[i, "th1"] - tra[i+1, "th1"]) > pi) {
      idxBrkTh1 <- c(idxBrkTh1, i+1)
    }
    if (abs(tra[i, "th2"] - tra[i+1, "th2"]) > pi) {
      idxBrkTh2 <- c(idxBrkTh2, i+1)
    }
  }
  # add the first point if necessary
  if (length(idxBrkTh1) == 0 || idxBrkTh1[1] != 1) {
    idxBrkTh1 <- c(1, idxBrkTh1)
    }
  if (length(idxBrkTh2) == 0 || idxBrkTh2[1] != 1) {
    idxBrkTh2 <- c(1, idxBrkTh2)
  }

  # add the last point if necessary
  len <- length(idxBrkTh1)
  if (idxBrkTh1[len] != nrow(tra)) { idxBrkTh1 <- c(idxBrkTh1, nrow(tra)) }
  len <- length(idxBrkTh2)
  if (idxBrkTh2[len] != nrow(tra)) { idxBrkTh2 <- c(idxBrkTh2, nrow(tra)) }

  # Done with the preparations. Now we can plot ...
  
  # First plot: theta over time
  t <- tra[, "time"]
  th1 <- tra[, "th1"]  # theta of first pendulum
  th2 <- tra[, "th2"]  # theta of second pendulum
  plot(t, th1,
       type = "n",
       ylim = c(min(c(th1, th2)), max(c(th1, th2))),
       xlab = "Time (sec)", ylab = "Theta 1 , Theta 2 (rad)",
       main = "Double Pendulum",
       col = myCols[1])
  for (i in 1:(length(idxBrkTh1) - 1)) {
    xi <- idxBrkTh1[i]
    xa <- idxBrkTh1[i+1] - 1
    points(t[xi:xa], th1[xi:xa], type = "l", lwd = lwdTra, col = th1Col)
  }
  for (i in 1:(length(idxBrkTh2) - 1)) {
    yi <- idxBrkTh2[i]
    ya <- idxBrkTh2[i+1] - 1
    points(t[yi:ya], th2[yi:ya], type = "l", lwd = lwdTra, col = th2Col)
  }
  legend("bottomright",
         legend = c("Theta 1", "Theta 2"),
         lty = 1,
         bty = "n",
         col = c(th1Col, th2Col))

  # Second plot: phase space
  th1 <- tra[, "th1"]
  th2 <- tra[, "th2"]
  plot(th1, th2,
       xlab = "Theta1", ylab = "Theta2",
       main = "Phase Space Plot of Double Pendulum",
       type = "n",
       col = myCols)
  # merge the breakpoints
  brkAll <- unique(sort(c(idxBrkTh1, idxBrkTh2)))
  for (i in 1:(length(brkAll) - 1)) {
    idxI <- brkAll[i]
    idxA <- brkAll[i+1] - 1
    plotrix::color.scale.lines(th1[idxI:idxA],
                               th2[idxI:idxA],
                               col = myCols[idxI:idxA],
                               lwd = lwdPhase)
  }

  legend("bottomright",
         legend = c("t = 0",
                    sprintf("t = %d",
                            round(tra[nrow(tra), "time"]))),
         fill =  c(myCols[1], myCols[length(myCols)]),
         bty = "n")

  # Third plot: Power Spectra of theta 1 and theta 2
  # x-axis points are frequencies (discretized to the stated bandwidth)
  # y-axis values are the "power" of a particular frequency in the time-series,
  #     i.e. how much a harmonic function of the frequency at the x-axis
  #     contributes to the time series data.
  sp1 <- spectrum(th1, log = "no", plot = FALSE)
  sp2 <- spectrum(th2, log = "no", plot = FALSE)
  plot(sp1$freq, sp1$spec,
       xlim=c(0, 100/length(th1)),
       ylim = c(0, max(c(sp1$spec, sp2$spec))),
       xlab = sprintf("frequency (bandwith = %.5f)", sp1$bandwidth),
       ylab = "spectrum",
       main = "Power Spectrum of Double Pendulum",
       type = "n")
  points(sp1$freq, sp1$spec, type ="l", lwd=3, col=th1Col)
  points(sp2$freq, sp2$spec, type="l", lwd=1.5, col=th2Col)
  legend("topright",
         legend = c("Theta 1", "Theta 2"),
         lty = 1,
         bty = "n",
         col = c(th1Col, th2Col))

  par(oPar)

}

```



## Compute trajectories


#### First case: stable

Small angles, both pendulums starting at positive angles. Inner pendulum long, outer pendulum short and twice as heavy as the inner one.

``` {r sync}
N <- 20  # (seconds)
times <- seq(0, N, length.out = 5000) # 5,000 timepoints from 0 to N seconds

# Initial state: theta (angle) and omega (angular velocity)
state <- c(th1 = deg2rad( 27), om1 = 0, th2 = deg2rad(41), om2 = 0)

# Pendulum parameters: lengths and masses
parm <- list(L1 = 2.4, m1 = 0.8, L2 = 0.9, m2 = 1.9)

# Solve this
myTrajectory <- deSolve::ode(y = state,
                       times = times,
                       func = doublePendulum,
                       parms = parm,
                       method = "rk4")

# Plot it (Open a new graphics window with dev.new() and make it large.)
plotPendulum(myTrajectory)
```

This trajectory looks pretty stable - you can try running it for a longer time ...

#### Second case:

Small angles, pendulums starting at opposite sides of the vertical. Lengths and masses are the same as for the first case.

``` {r async}
N <- 20
times <- seq(0, N, length.out = 5000)
state <- c(th1 = deg2rad( 333), om1 = 0, th2 = deg2rad(41), om2 = 0)
parm <- list(L1 = 2.4, m1 = 0.8, L2 = 0.9, m2 = 1.9)
myTrajectory <- deSolve::ode(y = state,  times = times, func = doublePendulum,
                       parms = parm, method = "rk4")
plotPendulum(myTrajectory)
```

The power spectrum shows clearly defined peaks around characteristic frequencies


#### Third case: Chaos

Same as previous, but pendulums raised to +- 17 degrees away from vertical up.

``` {r chaotic}
N <- 20
times <- seq(0, N, length.out = 5000)
state <- c(th1 = deg2rad( 193), om1 = 0, th2 = deg2rad(163), om2 = 0)
parm <- list(L1 = 2.4, m1 = 0.8, L2 = 0.9, m2 = 1.9)
myTrajectory <- deSolve::ode(y = state,  times = times, func = doublePendulum,
                       parms = parm, method = "rk4")
plotPendulum(myTrajectory)
```

This is chaotic behaviour. The dominant peaks in the power spectrum spread out into more and more subpeaks.


#### Fourth case:

Same as previous, but going for a longer time.

``` {r longChaos}
N <- 100  # (seconds)
times <- seq(0, N, length.out = 10000)
state <- c(th1 = deg2rad( 193), om1 = 0, th2 = deg2rad(163), om2 = 0)
parm <- list(L1 = 1.1, m1 = 0.9, L2 = 0.9, m2 = 1.1)
myTrajectory <- deSolve::ode(y = state,  times = times, func = doublePendulum,
                             parms = parm, method = "rk4")
plotPendulum(myTrajectory, lwdPhase = 1.0)
```

Look at the power spectrum!


#### The edge of chaos
... a domain that moves between ordered and chaotic regimes.

``` {r edgeOfChaos}
N <- 300  # (seconds)
times <- seq(0, N, length.out = 20000)
deg <- 68
state <- c(th1 = deg2rad(360-deg), om1 = 0, th2 = deg2rad(deg), om2 = 0)
parm <- list(L1 = 2.4, m1 = 0.8, L2 = 0.9, m2 = 1.9)
myTrajectory <- deSolve::ode(y = state,  times = times, func = doublePendulum,
                             parms = parm, method = "rk4")
plotPendulum(myTrajectory, lwdPhase = 1.0)
```

As the power spectrum shows, there is still structure in the angle-trajectories, but there are many peaks and especially for Theta 2 we can't really speak of a dominant peak around a characteristic frequency anymore.



That's all.






----

## Further Reading ##


<p>&nbsp;
<p>

----

<p>&nbsp;
<p>

<div class="meta">
#### Meta ####

Version 2024.1.5

<u>Versions:</u><br/>
<ul>
<li>1.5  Markdown version.</li>
<li>1.4  2024 Updates. Added Hénon map; improved code for double pendulum to wrap trajectory cleanly into [0,2pi]; prettier colors. Prepare to move code into R Markdown.</li>
<li>1.3  Fix faulty double pendulum code. Add better plotting: three panels in one window, power spectrum analysis, plot legends, explanations.</li>
<li>1.2  Annotate the logistic map operations more. Allow different lengths and masses for the coupled pendulums. Comment the code.</li>
<li>1.1  Generalize to other chaotic systems. Add Double Pendulum. Start the Neural Network Exploration.</li>
<li>1.0  Logistic map code</li>
</ul>
</div>


<!-- [END] -->
