---
title: "Random Phobias"
author: "Boris Steipe <boris.steipe@utoronto.ca>"
date: "2024-10-03"
output: 
  html_document:
#    number_sections: yes
#    toc: yes
---
<!--
This is a brief, practical introduction to R Markdown.

The text you are reading is enclosed in HTML comment tags, it will be
hidden in the HTML document that this page produces.

The document is an "R markdown" document (and that's is why it has the
extension ".Rmd"), and this superb technology allows us to freely mix
text and code, and share results. 

Introducing Markdown
====================
Markdown is a "lightweight" markup language designed to make writing and
formatting text simple and intuitive. At least simpler and more intuitive
than HTML. Markdown uses plain text syntax to create rich text elements
like headings, lists, links, and tables. For example, to make a word bold,
you simply wrap it in double asterisks like this: **bold**. It is not as 
versatile and powerful as a markup language like HTML, but it is easy to
learn and straightforward to read. Moreover, the resulting documents can
be exported to multiple formats, such as PDF, Word, and HTML.

Markdown and Literate Programming
=================================
Literate programming is the idea of weaving narrative text with code to
produce documents that are both human-readable and executable. And Markdown
is very suitable for this purpose. This concept is fully realized in
R Markdown, an extension of Markdown used in R. R Markdown allows you to
write a single document that contains R code chunks, Markdown-formatted
text, and outputs such as tables, plots, and formatted results. The 
R Markdown document (.Rmd) is then rendered using a toolchain that includes
the knitr and rmarkdown packages in R, which transform the document into a
final output format, such as HTML, PDF, or Word.

In practice, to create an R Markdown document in R Studio, you start by
creating a new .Rmd file and using the template that R Studio provides.
You then intersperse code blocks (enclosed in triple backticks) with
Markdown text. Finally, when you click the Knit button at the top of the
script pane, R Studio will execute the code, generate outputs, and merge
them with your text to produce a final, polished document. This workflow
makes R Markdown the perfect format for reproducible research, data analysis,
and reporting, as it ensures your work is documented in a way that others
can understand, replicate, and build upon.

There are two ways to work with this file:

A: Rendering to HTML
====================
Click on the "Knit" button at the top, to run the code contained in this file.
You will obtain a structured document in HTML format that opens in the
Web browser that is built into RStudio. 

B: Working Interactively
========================
You can change the code in the code "chunks" to try variations interactively.
In this case you have to have to (a) prepare the environment by clicking on the
"Run All Chunks Above" icon at the right-hand side of the chunk, (b) make your
changes, and (c) click on the green right-pointing triangle "Run Current Chunk"
to execute the changed code.

It is probably best to first render the document (Option A), study the
rendered result, and then go back to specific points for experimentation
(Option B) to understand the code better.

-->

```{css, echo = FALSE}
.striped tr:nth-child(even) {
  background: #eaf1ff;
}
.striped td, .striped th {
  padding: 5px;
}
.meta {
  font-size: 0.8em;
  color: #bbbbbb;
}
```
<small>Random Phobias - sample code to introduce R Markdown documents.
(c) Boris Steipe 2015-2024.</small>






```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache.path = 'tmp/')
```

## Phobias! ##
We all have some, but we could always use more. How to know them all? With this code we access the [Wikipedia list of phobias](https://en.wikipedia.org/wiki/List_of_phobias), scrape the contents and assemble a dataframe. Then we write a function to retrieve a random phobia, which we can subsequently ponder on - either to delight in the fact that we don't have that fear, or to add to our daily quota of anxieties <small>(like our well-founded [fear of bad programming practice](http://xkcd.com/292/))</small>.

To load the list, we will "screenscrape" the contents of Wikipedia's [List of Phobias](https://en.wikipedia.org/wiki/List_of_phobias). First, we install the `rvest` library and the `xml2` library from CRAN, if we don't have it.
```{r packages}
if (! requireNamespace("rvest", quietly=TRUE)) {
  install.packages("rvest")
}
if (! requireNamespace("xml2", quietly=TRUE)) {
  install.packages("xml2")
}
```

`xml2::` handles reading and parsing of documents. The `rvest::` package was designed for screenscraping and has functions to make our life very easy: it accesses the response of an `xml2::` query, looks for all HTML formatted tables, parses them with an XPATH expression and returns them as lists from which we can get data frames.

To keep our code explicit and maintainable, we don't use the `library()` function to load the package contents, but use the `package::function()` idiom instead:

```{r getPageData, cache=TRUE}
pageSource <- xml2::read_html("https://en.wikipedia.org/wiki/List_of_phobias")
allTables <- rvest::html_table(pageSource, fill = TRUE)
```

There are ```r length(allTables)``` tables in the list, but the ones we are interested in are data frames with two columns named `Phobia` and `Condition`.

```{r collateTables, cache=TRUE}
phobiaTable <- data.frame(Phobia = character(), Condition = character())
for (i in seq_along(allTables)) {
  df <- allTables[[i]]
  if (paste(colnames(df), collapse = " ") == "Phobia Condition") {
    phobiaTable <- rbind(phobiaTable, df)
  }
}
```

A bit of cleanup is required. 

**First**, we want to remove the Wikipedia References. They are formatted as numbers in square brackets and the function `gsub()` with an appropriate regular expression will get rid of them.

```{r removeReferences, cache=TRUE}

phobiaTable$Phobia <- gsub("\\[\\d+\\]", "", phobiaTable$Phobia)
phobiaTable$Condition <- gsub("\\[\\d+\\]", "", phobiaTable$Condition)
```

<small>Regexes can be confusing - here is how `"\\[\\d+\\]"` works:

* `\\[`: Matches a literal opening square bracket `[`.
* `\\d+`: Matches one or more digits.
* `\\]`: Matches a literal closing square bracket `]`.

Note that all three elements need to be "escaped", but for different reasons. We need to escape the square brackets because those have a special meaning in regular expressions (they collect "character classes"). But here we want to **remove** their special meaning, so we escape them. However "d" does not have a special meaning - it is just the literal character "d". In our case we have to **give** it a special meaning though, it should match a _digit_, i.e. one of the numerals in 0-9. And we **give** it a special meaning by escaping it. Special meaning characters in regexes are `\^$.| .*?+()[` and `{`. Regexes have a bit of a learning curve ...

And one more thing: in R, we need to double the escape character. R strings use the backslash to indicate special characters (e.g., `\n` for a newline, or `\"` for a literal quotation mark). We need to **remove** this special meaning and send the literal backslash to the regex engine and we do this by doubling the backslash (`\\`).

In short, **the first backslash is for R**, and the **second backslash is for the regex engine**.
</small>

**Second**, many of the phobias are listed with various synonyms, separated by commas. Let's keep only the first one of those, the `strsplit()` function will break them apart.

```{r removeSynonyms, cache=TRUE}
for (i in seq_along(phobiaTable$Phobia)) {
  phobiaTable$Phobia[i] <- strsplit(phobiaTable$Phobia[i], ", ")[[1]][1]
}
```
For each phobia, we split the string on the combo `, `, take the first item of the list that `strsplit()` returns - i.e. `[[1]]` and take only the first element from there.

Done, we collected ```r nrow(phobiaTable)``` phobias. Let's randomly select a few and print them.


<p>&nbsp;
<p>

```{r , ref.label="randRow", echo=FALSE}
```

**Table**: seven random phobias<br/>
```{r renderPhobiaTable, echo=FALSE, results='asis'}
sel <- sample(1:nrow(phobiaTable), 7)
knitr::kable(phobiaTable[sel, ], table.attr = "class=\"striped\"", format = "html")
```

<p>&nbsp;
<p>

## Phobia of the Day ##

To pick a single random phobia from the list, we take a (pseudo) random sample of size 1 from the number of rows in the `phobiaFrame` object. And we can set a "seed" for the random number generator that is derived from other information. Our function thus returns a random row from a matrix or dataframe, and it uses an additional argument: `seed`. This can either be created from todays date (default), or you can use any integer which is then passed to R's `set.seed()` function.

```{r randRow}
randRow <- function(M, seed = as.integer(format(Sys.Date(), "%y%m%d"))) {
  # Return a (pseudo) random row from a dataframe M.
  oldseed <- .Random.seed         # play nice: save the original RNG state ...
  set.seed(as.integer(seed))      # initialize the RNG
  r <- M[sample(1:nrow(M), 1), ]  # fetch one random row
  .Random.seed  <- oldseed        # ... restore the RNG state
  return(r)
}
```

<p>&nbsp;
<p>
With this useful tool we can ponder on our favourite phobia of the day. For today, let it be **`r randRow(phobiaTable)[2]`**, the `r randRow(phobiaTable)[1]`.

_`r randRow(phobiaTable)[1]`_! Really!!? Awful.

<p>&nbsp;
<p>

## The Long and Short of Phobias ##

Finally: let's plot a histogram of phobia name lengths just to illustrate how plots work in Rmarkdown documents. 

```{r preProcess}

# select only phobia names that end with the word "phobia" and that have no blank spaces in the name
sel <-   grepl(".phobia$", phobiaTable$Phobia) &
       ! grepl(" ",        phobiaTable$Phobia)
names <- phobiaTable$Phobia[sel]
```

We collected ```r length(names)``` names for phobias. Here is a histogram of their lengths.

```{r showHist}

x <- nchar(names)
pShort <- names[which(x == min(x))[1]]  # pull out the shortest name ...
pLong  <- names[which(x == max(x))[1]]  # ... and the longest name too.
hist(x,
     breaks = 20,
     xlim = c(0,30),
     xlab = "name",
     ylab = "counts",
     main = "Length of phobia-names",
     sub = sprintf("Shortest: %s (%d), Longest: %s (%d)",
                   pShort, nchar(pShort), pLong, nchar(pLong)),
     cex.sub = 0.8,
     col ="#aef5ee",
     )
abline(v = 6, col = "#A00050")
text(0, -0.5, "...", pos = 3)
text(1, -0.5, "p", pos = 3)
text(2, -0.5, "h", pos = 3)
text(3, -0.5, "o", pos = 3)
text(4, -0.5, "b", pos = 3)
text(5, -0.5, "i", pos = 3)
text(6, -0.5, "a", pos = 3)

```

That's all.

----

## Further Reading ##
* [Yihui Xie's R Markdown Book](https://bookdown.org/yihui/rmarkdown/)
    * Yihui Xie is one of the thought leaders in the community and his work on the knitr and Bookdown packages have tranformed the field. Plus Yihui is a strong champion of free (or open source) software.

### RStudio documentation ###
* [R Markdown](https://rmarkdown.rstudio.com/index.html)
* [Syntax Basics](https://rmarkdown.rstudio.com/authoring_basics.html)
* [R Markdown course](https://rmarkdown.rstudio.com/lesson-1.html)

<p>&nbsp;
<p>

----

<p>&nbsp;
<p>

<div class="meta">
#### Meta ####

* Version 2024.01.0
* Code derived in part from `RandomPhobia.Rmd` written for the ABC course units.

</div>


<!-- [END] -->
