---
title: "computingAminoAcidSimilarity"
author: "boris.steipe@utoronto.ca"
format: html
---


In this quarto document, we combine some ideas we discussed in class about amino acid properties, and how to construct a function that can compute values of amino acid similarity that we can use to evaluate the structure of a genetic code.

This document was rendered with version: {{< meta version >}}


# Initializations.

We first check the existence of required files, and whether your course direcrtory is "in a sane state". That goes on under the hood of this first code chunk.
```{r}
#| echo: false
if (!dir.exists("../../dat")) {
  stop("Data directory 'dat/' not found. Is this file located in \"src/R\"?")
}

# Download two Excel tables into the dat/ folder (if those don't exist yet) ...

if (! file.exists("../../dat/Breimann_2024_Supplementary_Table_1.xlsx")) {
  ghURL  <- paste0("https://raw.githubusercontent.com/hyginn/CSB195/main/dat/",
                   "Breimann_2024_Supplementary_Table_1.xlsx")
  myPath <- paste0("../../dat/",
                   "Breimann_2024_Supplementary_Table_1.xlsx")
  download.file(url = ghURL, destfile = myPath, mode = "wb")
}

if (! file.exists("../../dat/Breimann_2024_Supplementary_Table_3.xlsx")) {
  ghURL  <- paste0("https://raw.githubusercontent.com/hyginn/CSB195/main/dat/",
                   "Breimann_2024_Supplementary_Table_3.xlsx")
  myPath <- paste0("../../dat/",
                   "Breimann_2024_Supplementary_Table_3.xlsx")
  download.file(url = ghURL, destfile = myPath, mode = "wb")
}

# Download a precomputed amino acid feature spcae file
if (! file.exists("../../dat/aaFeatureSpace.4.0.Rds")) {
  ghURL  <- paste0("https://raw.githubusercontent.com/hyginn/CSB195/main/dat/",
                   "aaFeatureSpace.4.0.Rds")
  myPath <- paste0("../../dat/",
                   "aaFeatureSpace.4.0.Rds")
  download.file(url = ghURL, destfile = myPath, mode = "wb")
}

# Define a consistent set of amino acid colours
AACOLS <- c(G = "#B9C2CD", P = "#D4CF82", C = "#F1DD38", A = "#D2DF40",
            V = "#B4E149", I = "#96E351", L = "#78E65A", M = "#6DCB6E",
            F = "#63B182", W = "#599797", Y = "#4F7CAB", H = "#4562BF",
            R = "#3B48D4", K = "#5F6BD8", Q = "#838FDC", N = "#A8B3E0",
            T = "#AA90BA", S = "#AD6D95", D = "#B04B70", E = "#B3294B")

Aaa2A <- function(txt) {
  # bare-bones - no checking done
  AAmap <- c(Ala = "A", Arg = "R", Asn = "N",
             Asp = "D", Cys = "C", Gln = "Q",
             Glu = "E", Gly = "G", His = "H",
             Ile = "I", Leu = "L", Lys = "K",
             Met = "M", Phe = "F", Pro = "P",
             Ser = "S", Thr = "T", Trp = "W",
             Tyr = "Y", Val = "V")
  return(AAmap[txt])
}

```


# Preparing the data

First, we read the sheets we need from the downloaded, published excel files that were distributed with the Amino Acid Ontology paper. These are the _Supplementary Tables_ 1 and 3. 

I often use names like "mySomething" as variable names, to avoid conflicts with
existing names or reserved words in the language:

```{r}

  myScales <- readxl::read_excel("../../dat/Breimann_2024_Supplementary_Table_1.xlsx",
                                 sheet = "Normalized")

  myCats   <- readxl::read_excel("../../dat/Breimann_2024_Supplementary_Table_3.xlsx",
                                 sheet = "Scales")

  # Always examine the result and check for sanity
  myScales[1:3,1:5]
  
  str(myCats)

```

Confirm that the IDs in both tables are the same. This kind of validation
is important. Here we use the function `setequal()` since we don't assume
that the order matches, only the IDs themselves.

```{r}
  setequal(colnames(myScales)[-1], myCats$scale_id)  # must be TRUE
  any(duplicated(myCats$scale_id))                   # must be FALSE
```

Define a vector of amino acid labels explicitly. This is the standard ordering of amino acids in this code. Using this variable guarantees that we don't accidentally mix up ordering. (For example the order in the AAindex is not the same as in aaOntology...) 
```{r}
  (aaLabels <- myScales$AA)
```

# Build a table of Amino Acid scale types and values

Build a skeleton data structure from the metadata in myCats. We do this
explicitly, rather than relying on whatever structure the incoming
data has: **build from intent!** This ensures we control the column names
and order, and that all assumptions we have about the resulting data
are met.

```{r}
aaOntology <- data.frame(
    scaleID          = myCats$scale_id,
    category         = myCats$category,
    subcategory      = myCats$subcategory,
    scaleName        = myCats$scale_name,
    scaleDescription = myCats$scale_description)

  rownames(aaOntology) <- aaOntology$scaleID # this gives us a convenient
                                             # way to access each row by
                                             # its ID

  # Add amino acid columns, initialized to NA
  for (aa in aaLabels) {
    aaOntology[ , aa] <- NA
  }
```

Fill in values scale by scale. Selection is by subcategory name,
not by row position. This makes the code robust to reordering. Use this
as a general principle: match by semantics (meaning, content), not by
structure (row index, relative position).

```{r}
  for (sID in aaOntology$scaleID) {
    aaOntology[sID, aaLabels] <- myScales[[sID]]
  }
```

Final sanity check
```{r}
  #    Pick one random scale ID ...
  (testID <- sample(aaOntology$scaleID, 1))

  # Print the reconstructed row (AA values only)
  cat("\nReconstructed values for scale", testID, ":\n")
  aaOntology[testID, aaLabels]

  # Print the original column from myScales
  cat("\nOriginal values from myScales:\n")
  myScales[[testID]]

  # Confirm equality (the identical() functionis VERY strict,
  # TRUE means success)
  cat("\nAre they identical?\n")
  identical(
    as.numeric(aaOntology[testID, aaLabels]),
    myScales[[testID]]
  )
```

## Selecting categories of amino acid scales

Prepare a data frame of subcategory IDs and 0 / 1 selectors
We let the code write code, which we can then examine interactively
and select by hand. This is a reproducible way to scaffold manual
decisions: generate template code programmatically, then curate it by hand.

```{r}

  mySubCats <- data.frame(cat = aaOntology$category,
                          subcat = aaOntology$subcategory,
                          use = numeric(nrow(aaOntology)))
```

Remove duplicates: we only need one row per unique subcategory. The `! duplicated()` idiom is very useful to keep only unique values. This keeps only the first occurrence of each - which is fine, for this
purpose they are all equivalent.

```{r}
  mySubCats <- mySubCats[! duplicated(mySubCats$subcat), ]
```

The following code can be used to write code ... the repeated lines of assignments were created this way. Once it was run, I copied the result, pasted it into this script (below) and used it to select categories for inclusion.

```r
  for (i in 1:nrow(mySubCats)) {
    cat(sprintf("mySubCats$use[%d] <- 0    # %s : %s\n",
                i,
                mySubCats$cat[i],
                mySubCats$subcat[i]))
  }
```

Why do we need to select? Why can't we just use all categories?

We want to use the result to compute "amino acid similarity" in order
to evaluate the structure of the genetic code. Therefore we need to
exclude all scales that are in some way influenced by the genetic
code structure itself, rather than the inherent properties of the amino acid itself. If we are not careful about this point, we easily fall into a trap of _circular reasoning_.

In particular, things like codon counts, exchange probabilities,
propensities to be observed in specific structural context, or categories
that are not well defined or unclassified should be excluded.

The following section is deliberately long and explicit: it’s our
record of assumptions. This ensures future readers (including
ourselves) know exactly what we excluded, and how we interpreted
our rationale above.

```{r}
  mySubCats$use[ 1] <- 1    # ASA/Volume : Accessible surface area (ASA)
  mySubCats$use[ 2] <- 1    # ASA/Volume : Buried
  mySubCats$use[ 3] <- 1    # ASA/Volume : Hydrophobic ASA
  mySubCats$use[ 4] <- 1    # ASA/Volume : Partial specific volume
  mySubCats$use[ 5] <- 1    # ASA/Volume : Volume
  mySubCats$use[ 6] <- 0    # Composition : AA composition
  mySubCats$use[ 7] <- 0    # Composition : AA composition (surface)
  mySubCats$use[ 8] <- 0    # Composition : Membrane proteins (MPs)
  mySubCats$use[ 9] <- 0    # Composition : Mitochondrial proteins
  mySubCats$use[10] <- 0    # Composition : MPs (anchor)
  mySubCats$use[11] <- 0    # Composition : Unclassified (Composition)
  mySubCats$use[12] <- 0    # Conformation : Coil
  mySubCats$use[13] <- 0    # Conformation : Coil (C-term)
  mySubCats$use[14] <- 0    # Conformation : Coil (N-term)
  mySubCats$use[15] <- 0    # Conformation : Linker (>14 AA)
  mySubCats$use[16] <- 0    # Conformation : Linker (6-14 AA)
  mySubCats$use[17] <- 0    # Conformation : Unclassified (Conformation)
  mySubCats$use[18] <- 0    # Conformation : α-helix
  mySubCats$use[19] <- 0    # Conformation : α-helix (C-cap)
  mySubCats$use[20] <- 0    # Conformation : α-helix (C-term)
  mySubCats$use[21] <- 0    # Conformation : α-helix (C-term, out)
  mySubCats$use[22] <- 0    # Conformation : α-helix (left-handed)
  mySubCats$use[23] <- 0    # Conformation : α-helix (N-cap)
  mySubCats$use[24] <- 0    # Conformation : α-helix (N-term)
  mySubCats$use[25] <- 0    # Conformation : α-helix (N-term, out)
  mySubCats$use[26] <- 0    # Conformation : α-helix (α-proteins)
  mySubCats$use[27] <- 0    # Conformation : β/α-bridge
  mySubCats$use[28] <- 0    # Conformation : β-sheet
  mySubCats$use[29] <- 0    # Conformation : β-sheet (C-term)
  mySubCats$use[30] <- 0    # Conformation : β-sheet (N-term)
  mySubCats$use[31] <- 0    # Conformation : β-strand
  mySubCats$use[32] <- 0    # Conformation : β-turn
  mySubCats$use[33] <- 0    # Conformation : β-turn (C-term)
  mySubCats$use[34] <- 0    # Conformation : β-turn (N-term)
  mySubCats$use[35] <- 0    # Conformation : β-turn (TM helix)
  mySubCats$use[36] <- 0    # Conformation : π-helix
  mySubCats$use[37] <- 1    # Energy : Charge
  mySubCats$use[38] <- 1    # Energy : Charge (negative)
  mySubCats$use[39] <- 1    # Energy : Charge (positive)
  mySubCats$use[40] <- 1    # Energy : Electron-ion interaction pot.
  mySubCats$use[41] <- 1    # Energy : Entropy
  mySubCats$use[42] <- 1    # Energy : Free energy (folding)
  mySubCats$use[43] <- 1    # Energy : Free energy (unfolding)
  mySubCats$use[44] <- 1    # Energy : Isoelectric point
  mySubCats$use[45] <- 1    # Energy : Non-bonded energy
  mySubCats$use[46] <- 1    # Energy : Unclassified (Energy)
  mySubCats$use[47] <- 0    # Others : Mutability
  mySubCats$use[48] <- 0    # Others : PC 1
  mySubCats$use[49] <- 0    # Others : PC 2
  mySubCats$use[50] <- 0    # Others : PC 3
  mySubCats$use[51] <- 0    # Others : PC 4
  mySubCats$use[52] <- 0    # Others : PC 5
  mySubCats$use[53] <- 0    # Others : Unclassified (Others)
  mySubCats$use[54] <- 1    # Polarity : Amphiphilicity
  mySubCats$use[55] <- 1    # Polarity : Amphiphilicity (α-helix)
  mySubCats$use[56] <- 1    # Polarity : Hydrophilicity
  mySubCats$use[57] <- 1    # Polarity : Hydrophobicity
  mySubCats$use[58] <- 1    # Polarity : Hydrophobicity (interface)
  mySubCats$use[59] <- 1    # Polarity : Hydrophobicity (surrounding)
  mySubCats$use[60] <- 1    # Polarity : Unclassified (Polarity)
  mySubCats$use[61] <- 0    # Shape : Graph (1. eigenvalue)
  mySubCats$use[62] <- 0    # Shape : Graph (2. eigenvalue)
  mySubCats$use[63] <- 1    # Shape : Reduced distance
  mySubCats$use[64] <- 1    # Shape : Shape and Surface
  mySubCats$use[65] <- 1    # Shape : Side chain length
  mySubCats$use[66] <- 1    # Shape : Steric parameter
  mySubCats$use[67] <- 0    # Shape : Unclassified (Shape)
  mySubCats$use[68] <- 1    # Structure-Activity : Backbone-dynamics (-CH)
  mySubCats$use[69] <- 1    # Structure-Activity : Backbone-dynamics (-NH)
  mySubCats$use[70] <- 1    # Structure-Activity : Flexibility
  mySubCats$use[71] <- 1    # Structure-Activity : Flexibility (2 rigid neighbors)
  mySubCats$use[72] <- 1    # Structure-Activity : Stability
  mySubCats$use[73] <- 1    # Structure-Activity : Stability (helix-coil)
  mySubCats$use[74] <- 0    # Structure-Activity : Unclassified (Structure-Activity)
```

Finally, we define a selection vector that selects only the desired
subcategories in the aaOntology table.
  
```{r}
  sel <- logical(nrow(aaOntology))
  for (i in seq_len(nrow(aaOntology))) {
    idx <- which(mySubCats$subcat == aaOntology$subcategory[i])  # which()
    sel[i] <- (mySubCats$use[idx] == 1)  # This is a bit of idiomatic R style:
                                         # wrapping the expression in
                                         # parentheses returns the result,
                                         # which can then be assigned
  }
```

The above selection could be written both more "pedestrian", by
running a loop that checks every single row, or more idiomatic, with
this single expression:

```r
  sel <- aaOntology$subcategory %in% mySubCats$subcat[mySubCats$use == 1]
```

... which coding style is preferred? Prefer the version that is most explicit for YOU.


# Principal Component Analysis (PCA)

PCA extracts the uncorrelated dimensions (principal components, PCs)
along which the categories of observations in a large data set vary
most strongly.

In our context:
- Observations = the 20 amino acids - as categories
- Variables    = the selected scales (hundreds of them)

PCA gives us a reduced-dimensional "feature space" for amino acids:
- Each _axis_ (principal component, PC) is orthogonal and ordered
    by how much variance it explains.
- Each _amino acid_ is projected into this space as a point with
    new coordinates (its "scores").
- Each _original scale_ contributes to these PCs with a weight
    (the "loadings"), which we can use for interpretation.

The coordinates of the amino acids allow us to compute distances between
them - this becomes a computed "similarity".

Comparing the axes with the original data-sets allows us to interpret
them as e.g. hydrophobicity, polarity, size, etc.

## Computing the PCA

Computing PCAs in R can be done with a single command. We apply it to
the selected rows. Rememebr that the data columns of each amino acid
are named according to the aaLabels...

```{r}
   aaPCA <- prcomp(t(aaOntology[sel, aaLabels]), scale. = FALSE)

   # confirm the correct dimensions of aaPCA$x: should be 20 x 20
   dim(aaPCA$x)
```

Note: Normally we would scale each variable to unit variance, (PCA is
sensitive to the magnitude of its input data), but here we already
normalized, the various inputs, so we don't scale them again. The
parameter scale. = TRUE is the default, once again: be wary of hidden
behaviour. For example, in the case that some of your data is meaningfully
more "important" you might actually want the variance to be larger.

## Selecting "important" components

We get as many dimensions as the number of categories in the contributing
data-sets, but the dimensions are (a) orthogonal, i.e. un-correlated, and
(b) ordered by importance, i.e. by the amount of "variance" in the
original data that each of them explains.

We can now keep only the most important ones, and discard all others.
"Most important" could mean, those that together explain 95% of the
variance in the data.

```{r}
  myColorGradient <- colorRampPalette(c("#02b597", "#549097", "#edf5f5"))(20)

  cumulativeProportions <- cumsum(aaPCA$sdev^2) / sum(aaPCA$sdev^2)
  barplot(cumulativeProportions,
          main="Cumulative Proportion of Variance Explained",
          xlab="Principal Component",
          ylab="Cumulative Proportion",
          ylim=c(0, 1),
          border="#AAAAAA",
          col=myColorGradient)

  myCutoff <- 0.95 # we define a cutoff for "most important"
  abline(h=myCutoff, col="#AA0000", lwd=2, lty=2)  # mark importance threshold

  # How many PCs to retain? (first PC where we pass 95% of variance)
  numPCsToRetain <- which(cumulativeProportions >= myCutoff)[1]
  cat(sprintf("Retaining %i principal components (%3.0f%% cutoff ).\n",
              numPCsToRetain,
              myCutoff * 100))
```
## Constructing a Feature Space

Simply use only columns `1:numPCsToRetain` and assign that to the variable
name we want:

```{r}

  aaFeatureSpace <- aaPCA$x[ , 1:1:numPCsToRetain]

```


# Relationships between amino acids

A natural way to explore such relationships is a "heatmap". Here we compute a heatmap from the distances (note:  smaller distances mean greater similarity). Above we have defined a feature space. If we measure Euclidean distances of all pairs of positions of amino acids in that space, we get an overview of all similarities between amino acids. A heatmap is a commmon way to visualize that: bright spots mean amino acids are far apart, dark spots mean they’re close, and block structures identify groups. Notice how biochemically similar residues cluster together — that is our abstract space lining up with biology.


```{r}
  dmat <- as.matrix(dist(aaFeatureSpace))

  pinkPeaksPal <- colorRampPalette(c("#ff295f", "#f0297f",
                             "#99999b", "#cccccd"), bias = 1.4)
  heatmap(dmat,
          symm = TRUE,
          col = c("#888888", pinkPeaksPal(99)),
          na.rm = TRUE,
          main = "Amino Acid Similarity (PCA Feature Space)")
```

# Interpreting the Principal Components

Each of the original scales contributed a little to every dimension
of our feature space - but this does not mean that this is now
a purely abstract mathematical artefact. We can interpret the dimensions
by comparing them back to the original scales.

## Corelations between PCs and observed scales

For example, let's take the third-most important PC, the third column in
our `aaFeatureSpace` matrix, and calculate a coefficient of correlation
with each of the original scales. I write this as a completely
"pedestrian" for-loop. Why?

In R, the `cor()` function is "vectorised" and can take whole
matrices at once. That can look elegant, but it relies on a set of
implicit assumptions:
- Are rows or columns the "observations"?
- What happens if one argument is a vector, and the other is a
  matrix?
- Will `cor()` silently drop dimensions or recycle values?

For an expert user, those details can be managed. But for learners,
**"implicit" almost always means "unexpected"**. A one-liner with `cor()`
might be concise, but it is also brittle: change the orientation of
a table, and suddenly you get "incompatible dimensions" errors - or
worse: plausible numbers that are wrong, but hard to spot.

By contrast, the explicit loop below states exactly what we intend:
 "For each scale (row), compute its correlation with this PC."

This may be more pedestrian, but it is:
- Readable: the intent is obvious.
- Robust: no hidden conventions about rows vs. columns.
- Fast enough: looping over a few hundred scales × 20 amino acids
     is trivial for R.

For teaching, the clarity of intent matters more than saving one
line of code. In fact, this is a good example of the broader
principle: in data analysis, **explicit beats implicit**, especially
when code is read by humans (or AIs) who need to understand the
logic, not just the result.

 **Defensive coding: explicitly align amino acid types:**
 
We want the scores of amino acids along some Principal Component.
But: row order can sometimes change (e.g. one-letter and three-letter
codes are sorted differently) and all of a sudden, without noticing,
you are comparing values for Ala - A (Apples) with Arg - R (oRanges),
which is the kind of silent error that is as awful in its consequences
as it is hard to spot.

By indexing with the order defined in `aaLabels`, we guarantee the rows
to be in the same order as our amino acid labels. (Mind you, they
are already sorted that way, because this is how we constructed them.
But that is something we only _know_, not something we can _see_ at
this point in the code.)

```{r}
  thisPC <- aaFeatureSpace[aaLabels, 3] # define which PC to use
  myCors <- numeric(nrow(aaOntology))   # a vector to store the results
  names(myCors) <- aaOntology$scaleID   # named, to keep IDs with results

  for (i in seq_along(myCors)) {        # iterate over each row
    # cor() the PC with the aaLabel-ordered values, coerced into a vector
    myCors[i] <- cor(thisPC, as.numeric(aaOntology[i, aaLabels]))
  }
  
  # Print out the 10 most positive and most negative correlations
  head(sort(myCors, decreasing = TRUE), 10)   # strongest positive
  head(sort(myCors, decreasing = FALSE), 10)  # strongest negative
```
## Examine the dimensions

We can also plot two dimensions of the feature space as a scatterplot:

```{r, fig.width=6, fig.height=6}
  myDim1 <- 2
  myDim2 <- 5
  plot(aaFeatureSpace[ , myDim1], aaFeatureSpace[ , myDim2],
       asp = 1.0,
       type = "n",
       main = "PC scatterplot",
       xlab = sprintf("principal component %i", myDim1),
       ylab = sprintf("principal component %i", myDim2))
  text(aaFeatureSpace[, myDim1], aaFeatureSpace[,myDim2],
       rownames(aaFeatureSpace),
       col = AACOLS[rownames(aaFeatureSpace)],
       cex = 2.0)
```


## Summarize and Save

In summary our "Feature Space" for amino acids:

- ... takes all of the original data into account, each scale according to
     its contribution to the variance;
- ... recognizes that with 20 categories of observations (the 20 amino
     acids), there are only 19 independent dimensions of variance;
- ... defines those independent dimensions along which the contributions
     are orthogonal, i.e. uncorrelated;
- ... selects from those only the most "important" ones, i.e. those that
     jointly explain most of the variance;
- ... contains each amino acid as a unique "point", its position given by
     its coordinates - which are the values along the dimensions of the
     feature space.


Save the feature space for re-use using saveRDS() ... (see below)
```{r}
  myFN <- "../../dat/aaFeatureSpace.2025.Rds"
  if (! file.exists(myFN)) {
    saveRDS(aaFeatureSpace, file = "../../dat/aaFeatureSpace.2025.Rds")
  }
  # If we wish to reload later:
  # aaFeatureSpace <- readRDS("dat/aaFeatureSpace.2025.Rds")
```

Note: The pair `saveRDS()` / `readRDS()` save and _recreate_ single compressed
R objects. You can assign the "value" of `readRDS()` to a new variable, if
you want, or to the same variable name you used when creating the object.

In contrast the pair `save()` / `load()`,  _restores_ an R object to its
original name. This might silently overwrite an object that currently
exists in your workspace.


#  MILESTONE: Amino Acid Feature Space

At this point
- we have taken a large set of biophysical and statistical observations
  of amino acids;
- we have scaled them to be numerically comparable among each other;
- we have selected catgories of observations that we believe are independent
  of the genetic code itself, and thus can be used to evaluate the
  genetic code;
- we have used PCA to remove any sampling bias and correlations between
  the various sets of observations;
- we have constructed a "feature space" constructed from the most
  important "Principal Components" of the dataset, which represents
  the whole information contained in the original observations.
- we have defined the position of each amino acid in this space.


# Amino Acid Similarity - the Function `aaSim()`

We are now ready to define:

>   **The similarity between two amino acids is the Euclidian distance between their positions in an amino acid feature space.**

This maps the diverse properties of a pair of amino acids to a single number.

With that we are ready to write a function that computes similarity. I define
this function as a "closure" - an R specialty which allows us to package
code and data into a convenient unit. You can skip this section if you
are only interested in the result.

```{r}
  aaSimConstructor <- function() {

    # Value: a function that computes pairwise amino acid distances.

    # Note:

    # This function returns another function as its output. The returned
    # function is a so-called "closure" - a combination of data in its
    # environment, and code instructions. By doing this, the auxiliary data for
    # our function is created only once, when the function is defined, not every
    # time it is called. Yet, since the objects are defined locally, we do not
    # risk overwriting objects in our workspace when we define the function.

    # Note:
    #    The resulting function takes as its input two amino acid
    #    one-letter symbols and returns the Euclidian distance between
    #    the two vectors in the feature space defined in SPACEFILE.
    #    a1, a1: two letters from "ACDEFGHIKLMNPQRSTVWY*"
    #    Value:  distance

    # === Parameters ==============================

    # Re-create the Feature Space object:
    SPACEFILE <- "../../dat/aaFeatureSpace.4.0.Rds"
    if (! file.exists(SPACEFILE)) {
      GHurl  <- paste0("https://raw.githubusercontent.com/hyginn/CSB195/main/",
                         "aaFeatureSpace.4.0.Rds")
        download.file(url = GHurl, destfile = SPACEFILE, mode = "wb")
    }

    # Stop codon distance:
    # The distance of an amino acid to a stop codon is STOPDIST times the
    # maximum distance in the distance matrix.
    STOPDIST <- 1.5

    # Recreate the Feature Space
    AASPACE <- readRDS(SPACEFILE)

    # Compute a 21 x 21 matrix of Euclidian distances between any pair of vectors
    # in AASPACE.
    AADMAT <- matrix(numeric(21 * 21), nrow = 21)
    # Fill the first 20 x 20 values with amino acid pair distances
    for (i in 1:20) {
      for (j in 1:20) {
        AADMAT[i, j] <- sqrt(sum((AASPACE[i, ] - AASPACE[j, ])^2))
      }
    }

    # Define distance to stop codons: distance of stop codon "*" to
    # any other codon as STOPDIST times the maximum distance in the
    # distance matrix.

    stopDist <- STOPDIST * max(AADMAT)
    AADMAT[ 21, ] <- stopDist
    AADMAT[ , 21] <- stopDist
    AADMAT[21,21] <- 0

    rownames(AADMAT) <- c(rownames(AASPACE), "*")
    colnames(AADMAT) <- c(rownames(AASPACE), "*")

    # Define a function to return the pairwise distance between two points
    # in the distance space
    myFun <- function(a1, a2) {
      return(AADMAT[a1, a2])
    }
    attr(myFun, "AADMAT") <- AADMAT
    attr(myFun, "alphabet") <- rownames(AADMAT)

    return(myFun)  # return the function
  }

  aaSim <- aaSimConstructor()   # Define the function aaSim()
  rm(aaSimConstructor)          # The constructor function is no longer needed
```

## Test `aaSim()` and validate

```{r}
  # Our similarities induce a METRIC space
  aaSim("Q", "Q") #  IDENTITY: d(x,x) == 0: x == y
  any(attr(aaSim, "AADMAT") < 0)  # FALSE: POSITIVITY
  aaSim("F", "I") #  Similar amino acids: distance is small.
  aaSim("Q", "F") #  Dissimilar amino acids: distance is large.
  aaSim("F", "Q") #  SYMMETRY: d(x, y) == d(y, x)
  aaSim("Q", "*") #  Distance between any amino acid and a stopncodon is large.
  aaSim("*", "*") #  Two stop codons: distance is zero.
  
  aaSim("G", "C")
  aaSim("C", "W")
  aaSim("G", "W")
  aaSim("G", "C") + aaSim("C", "W") <= aaSim("G", "W") # TRIANGLE INEQUALITY
  
```

### Distinct Amino Acids

Let's do a quick check of which amino acid is the most distinct, and which
one is the most "plain vanilla" among the twenty. We develop this from
pseudocode:

```
For each amino acid
  For each of the 19 other amino acids
    Compute the distance of the amino acid pair
  Compute the mean
```

```{r}
  mySims <- numeric(20)  # create an empty 20-element vector of numbers
  names(mySims) <- aaLabels

  #> For each amino acid
  for (thisA in aaLabels) {
    for (otherA in aaLabels) {
      #>   For each of the 19 other amino acids
      if (thisA == otherA) {
        next
      }
      #>     Sum the distance of the amino acid pair
      mySims[thisA] <- mySims[thisA] + aaSim(thisA, otherA)
    }
    #>   Compute the mean
    mySims[thisA] <- mySims[thisA] / 19
  }

  (mySims <- sort(mySims) ) # sort the vector and show the results
```

The most "distinct" amino acid is "R" (arginine), the most average amino
acid is "T" (threonine).

Now plot this:

```{r}
  barplot(mySims,
          main = "Amino Acid Similarity",
          ylab = "Mean pairwise distances in feature space (AU)",
          ylim = c(0.0, 10),
          col = AACOLS[names(mySims)],
          cex.names = 0.5)
```


**... Done.**




<!-- [END] -->



