---
title: "CLT Demo"
subtitle: "The Central Limit Theorem in Action"
author: "Boris Steipe"
date: last-modified      
version: "v 1.0"
format:
  html:
    toc: true
    number-sections: true
    code-fold: false
    include-in-header:
      text:  |
        <style>
          .version-info {
            font-size: 70%;
            color: #BBBBBB;
            margin-top: -10px;
          }
        </style>
---

::: version-info
Version {{< meta version >}}, last updated {{< meta date >}}.
:::

<!-- 
versions:
  1.0 - Basic, first posted version
-->

This Quarto document demonstrates a document that is rendered to html in the `./docs` subdirectory, whereupon it becomes visible as a static webpage in the "site" that is associated with your GitHub repository - on `github.io`. To achieve this:

1. Let's assume your name is _Kronecker_, you are working on a GitHub repository called `poiema`, and you author and edit a Quarto document called  `IntegerCatechism.qmd`. However, you don't keep it in the usual location `./src/qmd`, but you place it into the `./docs` directory of your project, to publish it.
1. The `./docs/` directory contains a configuration file: `./docs/_quarto.yml` which contains a line:<br/>
`  output-dir: .`<br/>
This overrides the default configuration (which renders into `./out`) and directs Quarto to render your `IntegerCatechism.html` file and a folder containing its images and other assets into the `./docs/` directory.
1. After rendering, you _commit_ all changes in the `./docs/` directory to _git_, and push everything to your GitHub repository.
1. When you do this for the first time, you need to trigger the creation of your "site" through your repository's **Settings**. Access them from your main repository page, find the **Pages** heading in the left-hand menu, make sure that the **Build and deployment** is set to "Deploy from a branch", in the **Branch** menu select "main", and select "/docs" as the folder. Then click **Save**. It may take a minute or two until your `.html` is available. This setting is recorded, you only need to do this once.
1. GitHub finds the `.html` files contained in your repository's `/docs` and its subdirectories, and makes them accessible. In your case, this would be the URL `https://kronecker.github.io/poiema/IntegerCatechism.html`.

This is the process I followed here.

## The Central Limit Theorem (CLT)

Normal distributions occur everywhere in nature. Exactly why has only been proven in 1901, by Aleksandr Lyapunov who stated the CLT clearly and proved its justification:  

Let $X_1, X_2, \dots, X_n$ denote a statistical sample of size $n$ from a population with expected value (average) $\mu$ and finite positive variance $\sigma^2$, and let $\bar{X}_{n}$ denote the sample mean (which is itself a random variable). Then:

$$
(\bar{X}_n - \mu)\sqrt{n} \;\;\xrightarrow{d}\;\; \mathcal{N}(0, \sigma^2)
\quad \text{as } n \to \infty.
$$



::: {.callout-note}
**Intuition**  
This means: If we take many independent observations from the same population, then average them, the distribution of those averages will look more and more like a _bell curve_ - or "Gaussian Distribution" - as the sample size grows — no matter what the original population looked like (as long as it has a finite variance).
:::

One way to arrive at this, is through the accumulation of small errors, or fluctuations, the result of many interactions with a constantly changing context.


## Modelling the CLT by random fluctuations

Here is an example. We draw 5,000 samples from a uniform distribution, using the R function `runif()`, and plot that as a histogram. The values are uniformly distributed between 0 and 1, and the histogram bars all reach approximately the same height.

Then we add small fluctuations to each of the original values - something positive or negative, less than 5% ..., and since positive and negative perturbations are equally likely, they average out to about zero, when you look at the entire population. But that does not mean the original values all stay in place. If I do that 100 times, the histogram of the result starts getting peaked at ~0.5, and starts spreading out at the sides. I can repeat this, 200 times, 500 time, 1000 time more - every time plotting a new histogram (overlaid with slightly different colors, for clarity), and as a result, the histogram begins to approximate a normal distribution.

(Technically I am convoluting my uniform distribution with Gaussian noise.)

Here it is in code:

```{r}
#| cache: TRUE

set.seed(271828)

n <- 5000                     # number of points
x <- runif(n)                 # initial set of points: uniform distribution
xa <-   4                     # boundaries for the histogram: positive ...
xi <- -xa                     # ... and negative
brk <- seq(xi, xa, len=101)   # histogram bin limits

hist(x, 
     breaks = brk,
     col = "lightgray",
     main = "From uniform to normal via accumulated perturbations",
     xlab = "",
     ylab = "F",
     xlim = c(xi, xa),
     ylim = c(0, 500))

# Now add our perturbations: 100 random values, uniformly distributed
# between -0.05 and 0.05 added to each number in x. These are vectorized
# operations in R - they are fast. Then we overlay the new histogram.
x <- x + rowSums(matrix(runif(n*100, -0.05, 0.05), nrow=n))
hist(x[x>xi & x<xa], breaks = brk, col = "#FF00AA33", add=TRUE)

# Note: for display we clip the values to lie between the bounds we are
#       considering for our histogram: x[x>xi & x<xa]. If we don't
#       do that, sooner or later a value would stray outside the bounds
#       and hist() would throw a tantrum. My, my.


# We repeat this with 200 perturbations ...
x <- x + rowSums(matrix(runif(n*200, -0.05, 0.05), nrow=n))
hist(x[x>xi & x<xa], breaks = brk, col = "#FF009944", add=TRUE)

# ... 500 ...
x <- x + rowSums(matrix(runif(n*500, -0.05, 0.05), nrow=n))
hist(x[x>xi & x<xa], breaks = brk, col = "#FF005555", add=TRUE)

# ... and 1,000. That's our final histogram.
x <- x + rowSums(matrix(runif(n*1000, -0.05, 0.05), nrow=n))
hist(x[x>xi & x<xa], breaks = brk, col = "#FF001166", add=TRUE)
```

As you can see, the perturbations cause the distribution to become rounded, with a peak at the mean of the original distribution: at 0.5.  However these are the "same" 5,000 points we started with, we just added a few random positive and negative numbers to each of them.


## Normal Distribution and Perturbations

Are the 5,000 values now normally distributed?

```{r}

# assign the hist() output to myH, so we can get at the values with which
# the plot was constructed.
myH <- hist(x[x>xi & x<xa], breaks = brk, col = "#FF001166",
            freq = FALSE,
            main = "Histogram of Perturbed Values",
            xlab = "", ylab = "F",
            xlim = c(xi, xa))

points(myH$mids,  # using the histogram-bin midpoints as the x-values of
                  # our curve, and computing the y-values with dnorm()
       dnorm(myH$mids, mean = mean(x), sd = sd(x)),
       col = "#AA0000",
       type = "l",
       lwd = 2)
```

The `points()` function overlays points over a previous plot, the parameter
`type = "l"` connects the points with line segments. The function `dnorm()`
evaluates a normal distribution of the form 

$$
f(x) = \frac{1}{\sqrt{2\pi\sigma^2}}
       \exp\!\left( -\frac{(x - \mu)^2}{2\sigma^2} \right)
$$
... where $\mu$ is the mean of our perturbed values, and $\sigma$ is their standard deviation. Note that the curve is defined by these two numbers only, the 5,000 individual values that determine the heights of the histogram bars do not matter - this is not a curve fit. It turns out, that the actual points are correctly described by the mathematical relationship, without any further assumptions.

Francis Galton expressed this beautifully:

> I know of scarcely anything so apt to impress the imagination as the wonderful form of cosmic order expressed by the "Law of Frequency of Error". The law would have been personified by the Greeks and deified, if they had known of it. It reigns with serenity and in complete self-effacement, amidst the wildest confusion. The huger the mob, and the greater the apparent anarchy, the more perfect is its sway. **It is the supreme law of Unreason**. Whenever a large sample of chaotic elements are taken in hand and marshalled in the order of their magnitude, an unsuspected and most beautiful form of regularity proves to have been latent all along.

<small>Galton, Francis (1889). [_Natural Inheritance_](https://galton.org/books/natural-inheritance/index.html) p.66. (emphasis mine).</small>

<small>This quote, and other historical details can be found on the [CLTs Wikipedia page](https://en.wikipedia.org/wiki/Central_limit_theorem).</small>




That is all.

<!-- [END] -->  








